% -*- TeX-master: "sbml-level-2-version-3"; fill-column: 66 -*-
% $Id$
% $Source$
% ----------------------------------------------------------------

\section{A standard format for the \token{annotation} \changed{element}}
\label{sec:finney-novere}
\label{sec:annotation-standard}

This section describes the standard non-proprietary format for
\token{annotation} elements when (a) referring to controlled
vocabulary terms and database identifiers which define and
describe biological and biochemical entities; and (b) describing
the creator of a model and its modification history. Such a
structured format should facilitate the generation of models
compliant with the MIRIAM guidelines for model
curation~\citep{le_novere:2005}.

This format should \emph{not} be used to refer to SBO terms
(Section~\ref{sec:sboTerm}), because SBO defines terms about
mathematical modeling constructs and not the biological and
biochemical entities that the mathematics represent.

This format describes the form of one of the top level elements
that could reside in a given \token{annotation} structure. As such
this format is compliant with the constraints placed on the form
of annotation elements described in
Section~\ref{sec:annotation-use}.


\subsection{Motivation}

The SBML structures described elsewhere in this document do not
have any biochemical or biological semantics.  The format
described in this section provides a scheme for linking SBML
structures to external resources so that those structures can have
such semantics.  The motivation for the introduction of this
scheme is similar to that given for the introduction of
\token{sboTerm} however this scheme is significantly more
flexible.


\subsection{XML namespaces in the standard annotation}

This format uses a restricted form of Dublin
Core~\citep{DCMI:2003} and BioModels qualifier elements (see
\url{http://www.biomodels.net/index.php?s=Qualifiers}) embedded in
RDF~\citep{w3c:2004}. It uses a number of external XML standards
and associated XML namespaces.
Table~\ref{tab:namespaces-for-standard-annotation} lists these
namespaces and relevant documentation on those namespaces.  The
format constrains the order of elements in these namespaces beyond
the constraints defined in the standard definitions for those
namespaces.  For each standard listed, the format only uses a
subset of the possible syntax defined by the given standard.  Thus
it is possible for an \token{annotation} element to include XML
that is compliant with those external standards but is not
compliant with the format described here.  Parsers wishing to
support this format should be aware that a valid
\token{annotation} element may contain an \token{rdf:RDF} element
which is not compliant with the format described here.  A parser
should check that all aspects of the syntax defined here before
assuming that the contained data is encoded in the format.

\begin{table}[bh]
  \vspace*{3ex}
  \small
  \centering
  \begin{edtable}{tabular}{lll}
    \toprule
& Namespace URI & Definition Document \\
    \midrule
    \token{dc} & \uri{http://purl.org/dc/elements/1.1/} & \citep{powell:2003}\\
    \token{rdf} & \uri{http://www.w3.org/1999/02/22-rdf-syntax-ns\#} & \citep{w3c:2004b} \\
    \token{dcterms} & \uri{http://purl.org/dc/terms/} & \citep{kokkelink:2002}\\
    & & \citep{DCMIUB:2005} \\
    \token{vcard} & \uri{http://www.w3.org/2001/vcard-rdf/3.0\#} & \citep{iannella:2001} \\
    \token {bqbiol} & \uri{http://biomodels.net/biology-qualifiers/} \\
    \token {bqmodel} & \uri{http://biomodels.net/model-qualifiers/} \\
    \bottomrule
  \end{edtable}
  \vspace*{-0.95ex}
  \caption{The XML standards used in the SBML standard format for annotation.
  The namespace prefix are shown to indicate only the prefix used in the main text.}
  \label{tab:namespaces-for-standard-annotation}
\end{table}


% The default page break is too awkward here.
\clearpage

%=============================================================================
\subsection{General syntax for the standard annotation}
%=============================================================================

An outline of the format syntax is shown below.

\vspace*{0.1ex}

\begin{example}
<SBML_ELEMENT +++ metaid="SBML_META_ID" +++ >
  +++
  <annotation>
    +++
    <rdf:RDF
      xmlns:rdf="http://www.w3.org/1999/02/22-rdf-syntax-ns\#"
      xmlns:dc="http://purl.org/dc/elements/1.1/"
      xmlns:dcterm="http://purl.org/dc/terms/"
      xmlns:vcard="http://www.w3.org/2001/vcard-rdf/3.0\#"
      xmlns:bqbiol="http://biomodels.net/biology-qualifiers/"
      xmlns:bqmodel="http://biomodels.net/model-qualifiers/"
    >
      <rdf:Description rdf:about="#SBML_META_ID">
        [MODEL_HISTORY]
        <RELATION_ELEMENT>
          <rdf:Bag>
            <rdf:li rdf:resource="URI" />
            ...
          </rdf:Bag>
        </RELATION_ELEMENT>
        ...
      </rdf:Description>
      +++
    </rdf:RDF>
    +++
  </annotation>
  +++
</SBML_ELEMENT>
\end{example}

The above outline shows the order of the elements. The capitalized
identifiers refer to generic strings of a particular type:
\texttt{SBML\_ELEMENT} refers to any SBML element name that can
contain an \token{annotation} element; \texttt{SBML\_META\_ID} is
a XML \primtype{ID} string; \texttt{RELATION\_ELEMENT} refers to
element names in either the namespace
\uri{http://biomodels.net/biology-qualifiers/} or
\uri{http://biomodels.net/model-qualifiers/}; and \texttt{URI} is
a URI.  \texttt{[MODEL\_HISTORY]} refers to an optional section
described in Section~\ref{sec:model-history-annotation} which can
only be present within SBML model elements. `\texttt{+++}' is a
placeholder for either no content or valid XML syntax that is not
defined by the standard annotation scheme but is consistent with
the relevant standards for the enclosing elements. `\texttt{...}'
is a placeholder for zero or more elements of the same form as the
immediately preceding element.  The precise form of whitespace and
the XML namespace prefix definitions is not constrained however
the elements and attributes must be in the namespaces shown. The
rest of this section describes the format formally in English.

In this format the annotation of an element is located in a single
\token{rdf:RDF} element contained within an SBML
\token{annotation} element. The annotation element can contain
other elements in any order as described in 
Section~\ref{sec:annotation-use}.  The format described in this
section only defines the form of the \token{rdf:RDF} element. The
containing SBML \SBase element must have a \token{metaid} \changed{attribute}
value. (As this attribute is of the type \primtype{ID} its value
must unique to the entire SBML document.)

The first element of the \token{rdf:RDF} element must be an
\token{rdf:Description} element with an \token{rdf:about}
attribute. (This format doesn't define the form of subsequent
elements of the \token{rdf:RDF} element.) The value of the
\token{rdf:about} attribute must be of the form
\texttt{\#<string>} where the string component is equal to the
value of the \token{metaid} \changed{attribute} of the containing SBML element.

The \token{rdf:Description} element can contain only an optional
model history section (see
Section~\ref{sec:model-history-annotation}) followed by a sequence
of zero or more BioModels relation elements. The optional model
history section can only be present within an SBML \Model element.
The specific type of the relation elements will vary depending on
the relationship between the SBML component and referenced
information or resource.

Although Section~\ref{sec:qualified-dc-annotation} describes the
detailed semantics of each of the relation element types, the
content of these elements follows the same form.  The BioModels
qualifiers relation elements must only contain a single
\token{rdf:Bag} element which in turn must only contain one or
more \token{rdf:li} elements.  The \token{rdf:li} elements must
only have a \token{rdf:resource} attribute containing a URI
referring to an information resource (See
Section~\ref{sec:uri-in-annotation}).

Annotations in this format can be located at different depths
within a model component as is appropriate.

%=============================================================================
\subsection{Use of URIs}
\label{sec:uri-in-annotation}
%=============================================================================

The format represents a set of relationships between the SBML
element and the resources referred to by the contained
\token{rdf:resource} attribute values.  The BioModels relation
elements simply define the type of the relationship.

For instance a \Species element representing a protein could be
annotated with a reference to the database UniProt by the
\uri{http://www.uniprot.org/\#P12999} resource identifier,
identifying exactly the protein described by the \Species element.
This identifier maps to a unique entry in UniProt which is never
deleted from the database. In the case of UniProt, this is the
``accession'' of the entry. When the entry is merged with another
one, both ``accession'' are conserved. Similarly in a controlled
vocabulary resource, each term is associated with a perennial
identifier. The UniProt entry also possess an ``entry name'' (the
Swiss-Prot ``identifier''), a ``protein name'', ``synonyms'' etc.
Only the ``accession'' is perennial and should be used.

The value of a \texttt{rdf:resource} attribute is a URI that both
uniquely identifies the resource, and the data in the resource. In
this case the resource constraining the identifier precedes the
'\#' symbol and the term or database identifier follows the '\#'
symbol. In the above example, the resource
\uri{http://www.uniprot.org/} includes the entry P12999.

The value of the \texttt{rdf:resource} attribute is a URI, not a
URL; as such, a URI does not have to reference a physical web
object but simply identifies a controlled vocabulary term or
database object (a URI is a label that, in this case, just happens
to look like a URL). For instance, a true URL for an Internet
resource such as \uri{http://www.uniprot.org/entry/P12999} might
correspond to the URI \uri{http://www.uniprot.org/\#P12999}.

SBML does not specify how a parser is to interpret a URI. In the
case of a transformation into a physical URL, there could be
several solutions. For instance, the URI
\uri{http://www.geneontology.org/\#GO:0007268} can be translated
into:

\noindent \uri{http://www.ebi.ac.uk/ego/DisplayGoTerm?selected=GO:0007268}\\
\noindent \uri{http://www.godatabase.org/cgi-bin/amigo/go.cgi?view=details\&query=GO:0007268}\\
\noindent
\uri{http://www.informatics.jax.org/searches/GO.cgi?id=GO:0007268}

Similarly the URI \uri{http://www.ec-code.org/\#3.5.4.4}
can refer to:

\noindent \uri{http://www.ebi.ac.uk/intenz/query?cmd=SearchEC\&ec=3.5.4.4}\\
\noindent \uri{http://www.expasy.org/cgi-bin/nicezyme.pl?3.5.4.4}\\
\noindent \uri{http://www.chem.qmul.ac.uk/iubmb/enzyme/EC3/5/4/4.html}\\
\noindent \uri{http://www.genome.jp/dbget-bin/www\_bget?ec:3.5.4.4}\\
\noindent etc.

To enable interoperability, the community has agreed on an initial
set of standardized valid URI syntax rules which may be used
within the standard annotation format. This set of rules is not
part of the SBML standard but will grow independently from
specific SBML Levels and Versions. As the set changes a given URI
syntax rule will not be modified although the physical resource
associated with the rule may change. These URIs will always be
composed as \texttt{resource\#id}. The MIRIAM resources
\url{http://www.ebi.ac.uk/compneur-srv/miriam-main/} lists URI syntaxes and
possible physical links to controlled vocabulary and databases.
Each entry contains a list of SBML and relation elements in which
the given URI can be appropriately embedded. To enable consistent
and thus useful links to external resources, the URI syntax rule
set must have a consistent view of the concepts represented by the
different SBML elements for the purposes of this format.  For
example as the rule set is designed to link SBML biological and
biochemical resources the rule set assumes that a \Species element
represents the concept of a biochemical entity type rather than
mathematical symbol. The URI rule list will evolve with the
evolution of databases and resources. The annotation format
described in this section does not require a simple parser of this
format to access this list.


%=============================================================================
\subsection{Relation elements}
\label{sec:qualified-dc-annotation}
%=============================================================================

To enable the format to encode different types of relationships
between SBML elements and resources, different BioModel qualifier
elements are used to enclose a set of \token{rdf:li} elements. The
relation elements imply a specific relationship between the
enclosing SBML element and the resources referenced by the
\token{rdf:li} elements.

The detailed semantics (i.e. from the perspective of automatic
parser) of the relation elements is defined by the URI list at
\url{http://www.ebi.ac.uk/compneur-srv/miriam-main/}, and thus is outside the
scope of SBML. The URI list generally assumes that the biological
entity represented by the element is the concept linked to the
reference resource.

Several relation elements with a given tag, enclosed in the same
SBML element, each represent an alternative annotation to the SBML
element. For example two \class{bqbiol:hasPart} elements within a
\class{Species} SBML element represent two different sets of
references to the parts making up the the chemical entity
represented by the species. (The species is not made up of all the
entities represented by all the references combined).

The complete list of the qualifier elements in the Biomodels
qualifier namespaces is documented at
\url{http://www.biomodels.net/index.php?s=Qualifiers}. The list is
divided into two namespaces one for model qualifiers
\uri{http://biomodels.net/biology-qualifiers/} (prefix used here
\token{bqbiol}) and the other for biological qualifiers
\uri{http://biomodels.net/model-qualifiers/}) (prefix used here
\token{bqmodel}.  This list will only grow i.e no element will be
removed from the list. The following is the list of elements at
the time of writing:

%\begin{figure}[htb]
%  \vspace*{2pt}
%  \centering
%\[
%%\begin{CD}
%\textrm{SBML Element}  @>{qualifier}>> \textrm{Referenced Resource} \\
%@VV{represents}V                @VV{represents}V \\
%\textrm{Biological Entity A}   @>{relationship}>>
%\textrm{Biological Entity B}
%%\end{CD}
%\]
%  \caption{How the relationship between biological
%  entities are defined using relations between resources}
%  \label{fig:lenovere-annotation}
%\end{figure}

\begin{itemize}

\item \token{bqmodel:is} The modeling object encoded by the SBML
  component is the subject of the referenced resource. For
  instance, this qualifier might be used to link the model to a
  model database.

\item \token{bqmodel:isDescribedBy} The modeling object
  encoded by the SBML component is described by
  the referenced resource. This relation might be used to link
  SBML components to the literature that describes this model or
  this kinetic law.

\item \token{bqbiol:is} The biological entity represented by the
  SBML component is the subject of the referenced resource. This
  relation might be used to link a reaction to its exact
  counterpart in KEGG or Reactome for instance.

\item \token{bqbiol:hasPart} The biological entity represented by
  the SBML component includes the subject of the referenced
  resource, either physically or logically. This relation might be
  used to link a complex to the description of its components.

\item \token{bqbiol:isPartOf} The biological entity represented by
  the SBML component is a physical or logical part of the subject
  of the referenced resource. This relation might be used to link
  a component to the description of the complex is belongs to.

\item \token{bqbiol:isVersionOf} The biological entity represented
  by the SBML component is a version or an instance of the subject
  of the referenced resource.

\item \token{bqbiol:hasVersion} The subject of the referenced
  resource is a version or an instance of the biological entity
  represented by the SBML component.

\item \token{bqbiol:isHomologTo} The biological entity represented
  by the SBML component is homolog, to the subject of the
  referenced resource, i.e. they share a common ancestor.

\item \token{bqbiol:isDescribedBy} The biological entity
  represented by the SBML component is described by the referenced
  resource. This relation should be used for instance to link a
  species or a parameter to the literature that describes the
  quantity of the species or the value of the parameter.

\end{itemize}

%In all cases using the biology qualifiers, the `object' of the
%relation is the biological or biochemical object represented by
%the enclosing SBML element. In the cases of the model qualifiers,
%the `object' of the relation is the model component encoded by the
%enclosing SBML element. The resources referenced by the
%\class{rdf:li} elements contained within the relation element are
%the `object' of the relation.


%===============================================================
\subsection{Model history}
\label{sec:model-history-annotation}
%================================================================

When enclosed in an SBML \Model element, the format described in
previous sections can include additional elements to describe the
history of the model.  This history data must occur immediately
before the first BioModels relation elements.  These additional
elements encode information on the model creator and a sequence of
dates recording changes to the model. The syntax for this section
is outlined below.

\begin{example}
<dc:creator rdf:parseType="Resource">
  <rdf:Bag>
    <rdf:li rdf:parseType="Resource">
      [[
      +++
      <vCard:N rdf:parseType="Resource">
        <vCard:Family>FAMILY_NAME</vCard:Family>
        <vCard:Given>GIVEN_NAME</vCard:Given>
      </vCard:N>
      +++
      [<vCard:EMAIL>EMAIL_ADDRESS</vCard:EMAIL>]
      +++
      [<vCard:ORG>
        <vCard:Orgname>ORGANIZATION_NAME</vCard:Orgname>
      </vCard:ORG>]
      +++
      ]]
    </rdf:li>
    ...
  </rdf:Bag>
</dc:creator>
<dcterms:created rdf:parseType="Resource">
  <dcterms:W3CDTF>DATE</dcterms:W3CDTF>
</dcterms:created>
{<dcterms:modified rdf:parseType="Resource">
  <dcterms:W3CDTF>DATE</dcterms:W3CDTF>
</dcterms:modified>}
\end{example}

The order of elements is as shown above except that elements of
the format contained between \texttt{[[} and \texttt{]]} can occur
in any order.  The capitalized identifiers refer to generic
strings of a particular type: \texttt{FAMILY\_NAME} is the family
name of a person who created the model; \texttt{GIVEN\_NAME} is
the first name of the same person who created the model;
\texttt{EMAIL\_ADDRESS} is the email address of the same person
who created the model; and \texttt{ORGANIZATION\_NAME} is the name
of the organization with which the same person who created the
model is affiliated \texttt{DATE} is a date in W3C date
format~\citep{wolf:1998}. \texttt{W3CDTF}, \texttt{N},
\texttt{ORG} and \texttt{EMAIL} are literal strings. The elements
of the format contained between \texttt{[} and \texttt{]} are
optional. `\texttt{+++}' is a placeholder for either no content or
valid XML syntax that is not defined by the standard annotation
scheme but is consistent with the relevant standards for the
enclosing elements. `\texttt{...}' is a placeholder for zero or
more elements of the same form as the immediately preceding
element. The precise form of whitespace and the XML namespace
prefix definitions is not constrained.  The remaining text in this
section describes the syntax formally in English.

The additional elements of the model history sub-format consist in
sequence of a \token{dc:creator} element, a
\token{dcterms:created} element and zero or more
\token{dcterms:modified} elements.  All these elements must have
the attribute \token{rdf:parseType} set to \texttt{Resource}.

The \token{dc:creator} element describes the person who created
the SBML encoding of the model and contains a single
\token{rdf:Bag} element.  The \token{rdf:Bag} element can contain
any number of elements however the first element must be a
\token{rdf:li} element.  The \token{rdf:li} element can contain
any number of elements in any order.  The set of elements
contained with the \token{rdf:li} element can include the
following informative elements: \token{vCard:N},
\token{vCard:EMAIL} and \token{vCard:ORG}.  The \token{vCard:N}
contains the name of the creator and must consist of a sequence of
two elements: \token{vCard:Family} and the \token{vCard:Given}
whose content is the family (surname) and given (first) names of
the creator respectively.  The \token{vCard:N} must have the
attribute \token{rdf:parseType} set to \texttt{Resource}.  The
content of the \token{vCard:EMAIL} element must be the email
address of the creator.  The content of the \token{vCard:ORG}
element must contain a single \token{vCard:Orgname} element.  The
\token{vCard:Orgname} element must contain the name of an
organization to which the creator is affiliated.

The \token{dcterms:created} and \token{dcterms:modified} elements
must each contain a single \token{dcterms:W3CDTF} element whose
content is a date in W3C date format~\citep{wolf:1998} which is a
a profile of (restricted form of) ISO 8601.


%================================================================
\subsection{Examples}
%=================================================================

The following shows the annotation of a model with model creation
data and links to external resources:

\begin{example}
<model metaid="_180340" id="GMO" name="Goldbeter1991_MinMitOscil">
    <annotation>
        <rdf:RDF
                xmlns:rdf="http://www.w3.org/1999/02/22-rdf-syntax-ns\#"
                xmlns:dc="http://purl.org/dc/elements/1.1/"
                xmlns:dcterms="http://purl.org/dc/terms/"
                xmlns:vCard="http://www.w3.org/2001/vcard-rdf/3.0\#"
                xmlns:bqbiol="http://biomodels.net/biology-qualifiers/"
                xmlns:bqmodel="http://biomodels.net/model-qualifiers/"
        >
            <rdf:Description rdf:about="#_180340">
                <dc:creator rdf:parseType="Resource">
                    <rdf:Bag>
                        <rdf:li rdf:parseType="Resource">
                            <vCard:N rdf:parseType="Resource">
                                <vCard:Family>Shapiro</vCard:Family>
                                <vCard:Given>Bruce</vCard:Given>
                            </vCard:N>
                            <vCard:EMAIL>bshapiro@jpl.nasa.gov</vCard:EMAIL>
                            <vCard:ORG>
                                <vCard:Orgname>NASA Jet Propulsion Laboratory</vCard:Orgname>
                            </vCard:ORG>
                        </rdf:li>
                    </rdf:Bag>
                </dc:creator>
                <dcterms:created rdf:parseType="Resource">
                    <dcterms:W3CDTF>2005-02-06T23:39:40+00:00</dcterms:W3CDTF>
                </dcterms:created>
                <dcterms:modified rdf:parseType="Resource">
                    <dcterms:W3CDTF>2005-09-13T13:24:56+00:00</dcterms:W3CDTF>
                </dcterms:modified>
                <bqmodel:is>
                    <rdf:Bag>
                        <rdf:li rdf:resource="http://www.ebi.ac.uk/biomodels/#BIOMD0000000003"/>
                    </rdf:Bag>
                </bqmodel:is>
                <bqmodel:isDescribedBy>
                     <rdf:Bag>
                         <rdf:li rdf:resource="http://www.pubmed.gov/#1833774"/>
                     </rdf:Bag>
                </bqmodel:isDescribedBy>
                <bqbiol:isVersionOf>
                    <rdf:Bag>
                        <rdf:li rdf:resource="http://www.genome.jp/kegg/pathway/#hsa04110"/>
                        <rdf:li rdf:resource="http://www.reactome.org/#69278"/>
                    </rdf:Bag>
                </bqbiol:isVersionOf>
        </rdf:Description>
    </rdf:RDF>
</annotation>
\end{example}

The following example shows a \Reaction structure annotated
with a reference to its exact Reactome counterpart.

\begin{example}
<reaction id="cdc2Phospho" metaid="jb007">
  <annotation>
    <rdf:RDF
      xmlns:bqbiol="http://biomodels.net/biology-qualifiers/"
      xmlns:rdf="http://www.w3.org/1999/02/22-rdf-syntax-ns\#"
    >
      <rdf:Description rdf:about="#jb007">
        <bqbiol:is>
          <rdf:Bag>
            <rdf:li rdf:resource="http://www.reactome.org/\#170156"/>
          </rdf:Bag>
        </bqbiol:is>
      </rdf:Description>
    </rdf:RDF>
  </annotation>
  <listOfReactants>
    <speciesReference species="cdc2"/>
  </listOfReactants>
  <listOfProducts>
    <speciesReference species="cdc2-Y15P"/>
  </listOfProducts>
  <listOfModifiers>
    <modifierSpeciesReference species="wee1"/>
  </listOfModifiers>
</reaction>
\end{example}

The following example describes a species that represents a
complex between the protein calmodulin and calcium ions:

\begin{example}
<species id="Ca_calmodulin" metaid="cacam">
  <annotation>
    <rdf:RDF
      xmlns:rdf="http://www.w3.org/1999/02/22-rdf-syntax-ns\#"
      xmlns:bqbiol="http://biomodels.net/biology-qualifiers/"
    >
      <rdf:Description rdf:about="\#cacam">
        <bqbiol:hasPart>
          <rdf:Bag>
            <rdf:li rdf:resource="http://www.uniprot.org/\#P62158"/>
            <rdf:li rdf:resource="http://www.genome.jp/kegg/compound/\#C00076"/>
          </rdf:Bag>
        </bqbiol:hasPart>
      </rdf:Description>
    </rdf:RDF>
  </annotation>
</species>
\end{example}

The following example describes a species that represents either
``Calcium/calmodulin-dependent protein kinase type II alpha
chain'' or ``Calcium/calmodulin-dependent protein kinase type II
beta chain''. This is the case for instance in the somatic
cytoplasm of striatal medium-size spiny neurons, where both are
present but they cannot be functionally differentiated.

\begin{example}
<species id="calcium_calmodulin" metaid="cacam">
  <annotation>
    <rdf:RDF
      xmlns:rdf="http://www.w3.org/1999/02/22-rdf-syntax-ns\#"
      xmlns:bqbiol="http://biomodels.net/biology-qualifiers/"
    >
      <rdf:Description rdf:about="\#cacam">
        <bqbiol:hasVersion>
          <rdf:Bag>
            <rdf:li rdf:resource="http://www.uniprot.org/\#Q9UQM7"/>
            <rdf:li rdf:resource="http://www.uniprot.org/\#Q13554"/>
          </rdf:Bag>
        </bqbiol:hasVersion>
      </rdf:Description>
    </rdf:RDF>
  </annotation>
</species>
\end{example}

The above approach should not be used to describe ``any
Calcium/calmodulin-dependent protein kinase type II chain'',
because such an annotation requires referencing the products of
other genes such as gamma or delta. All the known proteins could
be enumerated, but such an approach would almost surely lead to
inaccuracies because biological knowledge continues to evolve.
Instead, the annotation should refer to generic information such
as Ensembl family ENSF00000000194 ``CALCIUM/CALMODULIN DEPENDENT
KINASE TYPE II CHAIN'' or PIR superfamily PIRSF000594
``Calcium/calmodulin-dependent protein kinase type~II''.

%While with \texttt{HasVersion}, the described component could
%represent several alternative, with \texttt{isVersionOf} the
%described component is one of the alternative understated by the
%referenced resource.

The following two examples show how to use the qualifier
\texttt{is Version Of}. The first example is the relationship
between a reaction and an EC code. An EC code describes an
enzymatic activity and an enzymatic reaction involving a
particular enzyme can be seen as an instance of this activity. For
instance the following reaction represents the phosphorylation of
a glutamate receptor by a complex calcium/calmodulin kinase II.

\begin{example}
<reaction id="NMDAR_phosphorylation" metaid="thx1138">
  <annotation>
    <rdf:RDF
      xmlns:bqbiol="http://biomodels.net/biology-qualifiers/"
      xmlns:rdf="http://www.w3.org/1999/02/22-rdf-syntax-ns\#"
    >
      <rdf:Description rdf:about="#thx1138">
        <bqbiol:isVersionOf>
          <rdf:Bag>
            <rdf:li rdf:resource="http://www.ec-code.org/\#2.7.1.17"/>
          </rdf:Bag>
        </bqbiol:isVersionOf>
      </rdf:Description>
    </rdf:RDF>
  </annotation>
  <listOfReactants>
    <speciesReference species="NMDAR"/>
  </listOfReactants>
  <listOfProducts>
    <speciesReference species="P-NMDAR"/>
  </listOfProducts>
  <listOfModifiers>
    <modifierSpeciesReference species="CaMKII"/>
  </listOfModifiers>
  <kineticLaw>
    <math xmlns="http://www.w3.org/1998/Math/MathML">
      <apply>
        <times/>
        <ci>CaMKII</ci>
        <ci>kcat</ci>
        <apply>
          <divide/>
          <ci>NMDAR</ci>
          <apply>
            </times>
            <ci>NMDAR</ci>
            <ci>Km</ci>
          </apply>
        </apply>
      </apply>
    </math>
    <listOfParameters>
      <parameter id="kcat" value="1"/>
      <parameter id="Km" value="5e-10"/>
    </listOfParameters>
  </kineticLaw>
</reaction>
\end{example}

The second example of the use of \class{isVersionOf} is the
complex between Calcium/calmodulin-dependent protein kinase type
II alpha chain and Calcium/calmodulin, that is only one of the
``calcium- and calmodulin-dependent protein kinase complexes''
described by the Gene Ontology term GO:0005954.

\begin{example}
<species id="CaCaMKII" metaid="C8H10N4O2">
  <annotation>
    <rdf:RDF
      xmlns:rdf="http://www.w3.org/1999/02/22-rdf-syntax-ns\#"
      xmlns:bqbiol="http://biomodels.net/biology-qualifiers/"
    >
      <rdf:Description rdf:about="\#C8H10N4O2">
        <bqbiol:isVersionOf>
          <rdf:Bag>
            <rdf:li rdf:resource="http://www.geneontology.org/\#GO:0005954"/>
          </rdf:Bag>
        </bqbiol:isVersionOf>
      </rdf:Description>
    </rdf:RDF>
  </annotation>
</species>
\end{example}

The previous case is different form the following one, although they
could seem similar at first sight. The
``Calcium/calmodulin-dependent protein kinase type II alpha
chain'' is a part of the above mentioned ``calcium- and
calmodulin-dependent protein kinase complex''.

\begin{example}
<species id="CaMKIIalpha" metaid="C10H14N2">
  <annotation>
    <rdf:RDF
      xmlns:rdf="http://www.w3.org/1999/02/22-rdf-syntax-ns\#"
      xmlns:bqbiol="http://biomodels.net/biology-qualifiers/"
    >
      <rdf:Description rdf:about="\#C10H14N2">
        <bqbiol:isPartOf>
          <rdf:Bag>
            <rdf:li rdf:resource="http://www.geneontology.org/\#GO:0005954"/>
          </rdf:Bag>
        </bqbiol:isPartOf>
      </rdf:Description>
    </rdf:RDF>
  </annotation>
</species>
\end{example}

It is possible describe a component with several alternative sets
of qualified annotations. For instance, the following species
represents a pool of  GMP, GDP and GTP. We annotate it with the
three corresponding KEGG compound identifiers but also with the
three corresponding ChEBI identifiers.  The two alternative
annotations are encoded in separate \token{hasVersion} qualifier
elements.

\begin{example}
<species id="GXP" metaid="GXP">
  <annotation>
    <rdf:RDF
      xmlns:rdf="http://www.w3.org/1999/02/22-rdf-syntax-ns\#"
      xmlns:bqbiol="http://biomodels.net/biology-qualifiers/"
    >
      <rdf:Description rdf:about="\#GXP">
        <bqbiol:hasVersion>
          <rdf:Bag>
            <rdf:li rdf:resource="http://www.ebi.ac.uk/chebi/\#CHEBI:17345"/>
            <rdf:li rdf:resource="http://www.ebi.ac.uk/chebi/\#CHEBI:17552"/>
            <rdf:li rdf:resource="http://www.ebi.ac.uk/chebi/\#CHEBI:17627"/>
          </rdf:Bag>
        </bqbiol:hasVersion>
        <bqbiol:hasVersion>
          <rdf:Bag>
            <rdf:li rdf:resource="http://www.genome.jp/kegg/compound/\#C00035"/>
            <rdf:li rdf:resource="http://www.genome.jp/kegg/compound/\#C00044"/>
            <rdf:li rdf:resource="http://www.genome.jp/kegg/compound/\#C00144"/>
          </rdf:Bag>
        </bqbiol:hasVersion>
      </rdf:Description>
    </rdf:RDF>
  </annotation>
</species>
\end{example}

The following example presents a reaction being actually the
combination of three different elementary molecular reactions. We
annotate it with the three corresponding KEGG reactions, but also
with the three corresponding enzymatic activities.  Again the two
\class{hasPart} elements represent two alternative annotations.
The process represented by the \class{Reaction} structure is
composed of three parts, and  not six parts.

\begin{example}
<reaction id="adenineProd" metaid="adeprod">
  <annotation>
    <rdf:RDF
      xmlns:bqbiol="http://biomodels.net/biology-qualifiers/"
      xmlns:rdf="http://www.w3.org/1999/02/22-rdf-syntax-ns\#"
    >
      <rdf:Description rdf:about="\#adeprod">
        <bqbiol:hasPart>
          <rdf:Bag>
            <rdf:li rdf:resource="http://www.ec-code.org/\#2.5.1.22"/>
            <rdf:li rdf:resource="http://www.ec-code.org/\#3.2.2.16"/>
            <rdf:li rdf:resource="http://www.ec-code.org/\#4.1.1.50"/>
          </rdf:Bag>
        </bqbiol:hasPart>
        <bqbiol:hasPart>
          <rdf:Bag>
            <rdf:li rdf:resource="http://www.genome.jp/kegg/reaction/\#R00178"/>
            <rdf:li rdf:resource="http://www.genome.jp/kegg/reaction/\#R01401"/>
            <rdf:li rdf:resource="http://www.genome.jp/kegg/reaction/\#R02869"/>
          </rdf:Bag>
        </bqbiol:hasPart>
      </rdf:Description>
    </rdf:RDF>
  </annotation>
</reaction>
\end{example}

It is possible to mix different URIs in a given set. The
following example presents two alternative annotations of the human
hemoglobin, the first with ChEBI heme and the second with KEGG
heme.

\begin{example}
<species id="heme" metaid="heme">
  <annotation>
    <rdf:RDF
      xmlns:rdf="http://www.w3.org/1999/02/22-rdf-syntax-ns\#"
      xmlns:bqbiol="http://biomodels.net/biology-qualifiers/"
    >
     <rdf:Description rdf:about="\#heme">
       <bqbiol:hasPart>
         <rdf:Bag>
           <rdf:li rdf:resource="http://www.uniprot.org/\#P69905"/>
           <rdf:li rdf:resource="http://www.uniprot.org/\#P68871"/>
           <rdf:li rdf:resource="http://www.ebi.ac.uk/chebi/\#CHEBI:17627">
         </rdf:Bag>
       </bqbiol:hasPart>
       <bqbiol:hasPart>
         <rdf:Bag>
          <rdf:li rdf:resource="http://www.uniprot.org/\#P69905"/>
           <rdf:li rdf:resource="http://www.uniprot.org/\#P68871"/>
           <rdf:li rdf:resource="http://www.genome.jp/kegg/compound/\#C00032"/>
         </rdf:Bag>
       </bqbiol:hasPart>
     </rdf:Description>
   </rdf:RDF>
  </annotation>
</species>
\end{example}

As formally defined above it is possible to use different
qualifiers in the same annotation element. The following
phosphorylation is annotated by its exact KEGG counterpart and by
the generic GO term ``phosphorylation''.

\begin{example}
<reaction id="phosphorylation" metaid="phosphorylation">
  <annotation>
    <rdf:RDF
      xmlns:bqbiol="http://biomodels.net/biology-qualifiers/"
      xmlns:rdf="http://www.w3.org/1999/02/22-rdf-syntax-ns\#"
    >
      <rdf:Description rdf:about="\#phosphorylation">
        <bqbiol:is>
          <rdf:Bag>
            <rdf:li rdf:resource="http://www.genome.jp/kegg/reaction/\#R03313" />
          </rdf:Bag>
        </bqbiol:is>
        <bqbiol:isVersionOf>
          <rdf:Bag>
            <rdf:li rdf:resource="http://www.geneontology.org/\#GO:0016310" />
          </rdf:Bag>
        </bqbiol:isVersionOf>
      </rdf:Description>
    </rdf:RDF>
  </annotation>
</reaction>
\end{example}
