% -*- TeX-master: "sbml-level-2-version-3"; fill-column: 66 -*-
% $Id$
% $Source$
% ----------------------------------------------------------------

\section{Example models expressed in XML using SBML}
\label{sec:xml-rep}
\label{sec:examples}

In this section, we present several examples of complete models
encoded in XML using SBML Level~2.

%Our approach to translating
%the UML-based structure definitions presented in the previous
%sections is described elsewhere~\citep{hucka:2000b}.
%Appendix~\ref{apdx:schemas} gives the full listing of an XML
%Schema corresponding to SBML Level~2.


%-----------------------------------------------------------------------------
\subsection{A simple example application of SBML}
\label{sec:modeleg}
%-----------------------------------------------------------------------------

Consider the following representation of an enzymatic reaction:
\begin{linenomath}
\begin{reaction}\centering\notag
  E + S \eqbm^{\cit k\sub{on}}_{\cit k\sub{off}} ES \yields^{\cit k\sub{cat}} E + P
\end{reaction}
\end{linenomath}
The following is the minimal SBML document encoding the model shown above:

\sbmlexample{enzymekinetics.xml}

In this example, the model has the identifier
\val{EnzymaticReaction}.  The model contains one compartment (with
identifier \val{cytosol}), four species (with identifiers
\val{ES}, \val{P}, \val{S}, and \val{E}), and two reactions
(\val{veq} and \val{vcat}).  The elements in the
\token{listOfReactants} and \token{listOfProducts} in each
reaction refer to the names of elements listed in the
\token{listOfSpecies}.  The correspondences between the various
elements is explicitly stated by the \token{speciesReference}
elements.

The model also features local parameter definitions in each
reaction.  In this case, the three parameters (\val{kon},
\val{koff}, \val{kcat}) all have unique identifiers and they could
also have just as easily been declared global parameters in the
model.  Local parameters frequently become more useful in larger
models, where it may become tedious to assign unique identifiers
for all the different parameters.


%-----------------------------------------------------------------------------
\subsection{Example involving units}
\label{apdx:units-eg}
%-----------------------------------------------------------------------------

The following model uses the units features of SBML Level~2. In
this model, the default value of \token{substance} is changed to
be mole units with a scale factor of $-3$, or millimoles.  This
sets the default substance units in the model.  The \token{size}
and \token{time} built-ins are left to their defaults, meaning
size is in litres and time is in seconds.  The result is that, in
this model, kinetic law formulas define rates in millimoles per
second and the species identifiers in them represent concentration
values in millimoles per litres.  All the \token{species} elements
set the initial amount of every given species to 1 millimole.  The
parameters \val{vm} and \val{km} are defined to be in millimoles
per litres per second, and milliMoles per litres, respectively.

\sbmlexample{units.xml}


%-----------------------------------------------------------------------------
\subsection{Example of a discrete version of a simple dimerization reaction}
\label{sec:discrete-eg}
%-----------------------------------------------------------------------------

\emph{(Model contributed by Darren J. Wilkinson, Newcastle
  University, Newcastle upon Tyne, UK.)}

This example illustrates subtle differences between models
formulated for use in a continuous simulation framework (\eg using
differential equations) and those intended for a discrete
simulation framework.  The model shown here is suitable for use
with a discrete stochastic simulation algorithm of the sort
developed by \cite{gillespie:1977}.  In such an approach, species
are described in terms of molecular counts and stimulation
proceeds by computing the probability of the time and identity of
the next reaction, then updating the species amounts
appropriately.

The model involves a simple dimerization reaction for a protein
named \val{P}:
\begin{linenomath}
\begin{equation*}
    2 P  \leftrightarrow  P_2
\end{equation*}
\end{linenomath}
The SBML representation is shown below.  There are several
important points to note.  First, the species \val{P} and \val{P2}
declare they are always in discrete amounts by using the flag
\token{hasOnlySubstanceUnits}=\val{true}.  This indicates that
when the species identifiers appear in mathematical formulas, the
units are \quantity{substance}, not the default of
\quantity{substance}/\quantity{size}.  A second point is that, as
a result, the corresponding ``kinetic law'' formulas do not need
volume corrections.  In Gillespie's approach, the constants in the
rate expressions (here, \val{c1} and \val{c2}) contain a
contribution from the kinetic constants of the reaction and the
size of the compartment in which the reactions take place.
Finally, it is worth noting the rate expression for the forward
reaction is a second-order mass-action reaction, but it is the
\emph{discrete} formulation of such a reaction
rate~\citep{gillespie:1977}.

\sbmlexample{dimerization.xml}

This example also illustrates the need to provide additional
information in a model so that software tools using different
mathematical frameworks can properly interpret it.  In this case,
a simulation tool designed for continuous ODE-based simulation
would likely misinterpret the model (in particular the reaction
rate formulas), unless it deduced that a discrete stochastic
simulation was intended.  One of the purposes of SBO annotations
(Section~\ref{sec:sboTerm}) is to enable such interpretation
without the need for deduction.


%-----------------------------------------------------------------------------
\subsection{Example involving assignment rules}
\label{apdx:rules-eg}
%-----------------------------------------------------------------------------

This section contains a model that simulates a system containing a
fast reaction.  This model uses rules to express the mathematics
of the fast reaction explicitly rather than using the \token{fast}
field on a reaction element.  The system modeled is
\begin{linenomath}
\begin{equation*}
  \begin{array}{@{}ccc@{}}
    X_0 & \overset{\underrightarrow{k_1 X_0}}{}           & S_1 \\[6pt]
    S_1 & \overset{\underrightarrow{k_f S_1 - k_r S_2}}{} & S_2 \\[6pt]
    S_2 & \overset{\underrightarrow{k_2 S_2}}{}           & X_1 \\[6pt]
  \end{array}
\end{equation*}
\begin{equation*}
    k_1 = 0.1, \quad k_2 = 0.15, \quad k_f = K_{eq} 10000, \quad k_r = 10000, \quad K_{eq} = 2.5.
\end{equation*}
\end{linenomath}
where $X_0$, $S_1$, and $S_2$ are species in concentration units,
and $k_1$, $k_2$, $k_f$, $k_r$, and $K_{eq}$ are parameters.  This
system of reactions can be approximated with the following new
system:
\begin{linenomath}
\begin{equation*}
  \begin{array}{@{}ccc@{}}
    X_0 & \overset{\underrightarrow{k_1 X_0}}{} & T \\[6pt]
    T & \overset{\underrightarrow{k_2 S_2}}{} & X_1\\[6pt]
  \end{array}
\end{equation*}
\begin{equation*}
\begin{aligned}
    S_1 &= \dfrac{T}{1 + K_{eq}} \\[6pt]
    S_2 &= K_{eq} S_1
  \end{aligned}
\end{equation*}
\end{linenomath}

where $T$ is a new species.  The following example SBML model
encodes the second system.

\sbmlexample{assignmentrules.xml}


%-----------------------------------------------------------------------------
\subsection{Example involving algebraic rules}
\label{sec:algeraiceg}
%-----------------------------------------------------------------------------

This section contains an example model that contains an
\AlgebraicRule structure.  The model contains a different
formulation of the fast reaction described in
Section~\ref{apdx:rules-eg}.  The system described in
Section~\ref{apdx:rules-eg} can be approximated with the following
system:
\begin{linenomath}
\begin{equation*}
  \begin{array}{@{}ccc@{}}
    X_0 & \overset{\underrightarrow{k_1 X_0}}{} & T \\[6pt]
    T & \overset{\underrightarrow{k_2 S_1}}{} & X_1
  \end{array}
\end{equation*}
\begin{equation*}
    S_2 = K_{eq} S_1\\[10pt]
\end{equation*}
\end{linenomath}
with the constraint:
\begin{linenomath}
\begin{equation*}
    S_1 + S_2 - T = 0
\end{equation*}
\end{linenomath}

The following example SBML model encodes this approximate form.

\sbmlexample{algebraicrules.xml}


%-----------------------------------------------------------------------------
\subsection{Example with combinations of
  \token{boundaryCondition} and \token{constant} values on \class{Species}
  with \class{RateRule} structures}
\label{sec:constantspecieseg}
%-----------------------------------------------------------------------------

In this section, we discuss a model that includes four species,
each with a different combination of values for their
\token{boundaryCondition} and \token{constant} fields.  The model
represents a hypothetical system containing one reaction,
\begin{linenomath}
\begin{equation*}
  \begin{array}{@{}ccc@{}}
    S_1 + S_2 & \overset{\underrightarrow{k_1 S_1 S_2 S_3}}{} & S_4 \\ \\[-4pt]
  \end{array}
\end{equation*}
\end{linenomath}
where $S_3$ is a species that catalyzes the conversion of species
$S_1$ and $S_2$ into $S_4$.  $S_1$ and $S_2$ are on the boundary
of the system (\ie $S_1$ and $S_2$ are reactants but their values
are not determined by a kinetic law).  The value of $S_1$ in the
system is determined over time by the rate rule:
\begin{linenomath}
\begin{equation*}
  \dfrac{d S_1}{d t} = k_2
\end{equation*}
\end{linenomath}
The values of constant parameters in the system are:
\begin{linenomath}
\begin{equation*}
    S_2 = 1, \quad S_3 = 2, \quad k_1 = 0.5, \quad k_2 = 0.1
  \end{equation*}
\end{linenomath}
and the initial values of species are:
\begin{linenomath}
\begin{equation*}
    S_1 = 0, \quad S_4 = 0
\end{equation*}
\end{linenomath}

The value of $S_1$ varies over time so in SBML $S_1$ has a
\token{constant} field with a default value of \val{false}.  The
values of $S_2$ and $S_3$ are fixed so in SBML they have a
\token{constant} field values of \val{true}.  $S_3$ only occurs as
a modifier so the value of its \token{boundaryCondition} field can
default to \val{false}.  $S_4$ is a product whose value is
determined by a kinetic law and therefore in the SBML
representation has \val{false} (the default) for both its
\token{boundaryCondition} and \token{constant} fields.

The following is the SBML rendition of the model shown above:

\sbmlexample{boundarycondition.xml}


%-----------------------------------------------------------------------------
\subsection{Example of translation from a multi-compartmental model to ODEs}
\label{sec:odeeg}
%-----------------------------------------------------------------------------

This section contains a model with 2 compartments and 4 reactions.
The model is derived from Lotka-Volterra, with the addition of a
reversible transport step.  When observed in a time-course
simulation, three of this model's species display damped
oscillations.

\begin{figure}[htb]
  \vspace*{5pt}
  \centering
  \begin{picture}(260,60)
    \put(0,10){\framebox(255,50)[tl]{ cytosol}}
    \put(10,19){\framebox(105,29)[tl]{ nucleus}}
    \put(24,26){$
        X + Y_{1n} \yields^{\cit k\sub{1}} 2\,Y_{1n}
        \eqbm^{\cit K\sub{T}} 2\,Y_{1c} + 2\,Y_2
        \yields^{\cit k\sub{2}} 4\,Y_2 \yields^{\cit k\sub{3}} \emptyset
        $}
  \end{picture}
  \vspace*{-8pt}
  \caption{A example multi-compartmental model.}
  \label{fig:multicomp}
\end{figure}

Figure~\ref{fig:multicomp} illustrates the arrangement of
compartments and reactions in the model
\token{LotkaVolterra\_tranport}.  The text of the SBML
representation of the model is shown below, and it is followed by
its complete translation into ordinary differential equations.  In
this SBML model, the reaction equations are in substance per time
units.  The reactions have also been simplified to reduce common
stoichiometric factors.  The species variables are in
concentration units; their initial quantities are declared using
the attribute \token{initialAmount} on the \token{species}
definitions, but since the attribute \token{hasOnlySubstanceUnits}
is \emph{not} set to true, the identifiers of the species
represent their concentrations when those identifiers appear in
mathematical expressions elsewhere in the model.  Note that the
species whose identifier is \val{X} is a boundary condition, as
indicated by the attribute \token{boundaryCondition}=\val{true} in
its definition.  The attribute \token{speciesType}=\val{Y} in the
definitions of \val{Y1n} and \val{Y1c} indicates that these
species are pools of the same participant, but located in
different compartments.

\sbmlexample{multicomp.xml}

The ODE translation of this model is as follows.  First, we give
the values of the constant parameters:
\begin{linenomath}
\begin{equation*}
  k_1 = 2500, \quad k_2 = 2500, \quad K_3 = 25000, \quad K_T = 2500
\end{equation*}
\end{linenomath}
Next, here are the initial conditions of the variables.  The
species variables $X$, $Y_{1n}$, $Y_{1c}$, and $Y_2$ in the following
equations are all in terms of concentrations, and we use $V_n$ to
represent for the size of compartment \val{nucleus} and $V_c$ to
represent for the size of compartment \val{cytoplasm}:
\begin{linenomath}
\begin{equation*}
  V_n = 1, \quad V_c = 5, \quad X = 1, \quad Y_{1n} = 1, \quad Y_{1c} = 0, \quad Y_2 = 1
\end{equation*}
\end{linenomath}
And finally, here are the differential equations:
\begin{linenomath}
\begin{align*}
  \dfrac{d X}{d t}    &= 0 \\[6pt]
  V_n \dfrac{d Y_{1n}}{d t} &= k_1 \cdot X \cdot Y_{1n} \cdot V_n - K_T \cdot (Y_{1n} - Y_{1c}) \cdot V_c
    && \text{reactions production and transport} \\[6pt]
  V_c \dfrac{d Y_{1c}}{d t} &= K_T \cdot (Y_{1n} - Y_{1c}) \cdot V_c - k_2 \cdot Y_{1c} \cdot Y_2 \cdot V_c
    && \text{reactions transport and transformation} \\[6pt]
  V_c \dfrac{d Y_2}{d t}   &= k_2 \cdot Y_{1c} \cdot Y_2 \cdot V_c - k_3 \cdot Y_2 \cdot V_c
    && \text{reactions transformation and degradation}
\end{align*}
\end{linenomath}

As formulated here, this example assumes constant volumes.  If the
sizes of the compartments \val{cytoplasm} or \val{nucleus} could
change during simulation, then it would be preferable to use a
different approach to constructing the differential equations.  In
this alternative approach, the ODEs would compute substance change
rather than concentration change, and the concentration values
would be computed using separate equations.  This approach is used
in Section~\ref{sec:about-kinetic-laws}.


%-----------------------------------------------------------------------------
\subsection{Example involving function definitions}
\label{sec:functioneg}
%-----------------------------------------------------------------------------

This section contains a model that uses the function definition
feature of SBML.  Consider the following hypothetical system:
\begin{linenomath}
\begin{equation*}
  \begin{array}{@{}ccc@{}}
    S_1 & \overset{\underrightarrow{f(S_1)}}{} & S_2 \\ \\[-4pt]
  \end{array}
\end{equation*}
\end{linenomath}
where
\begin{linenomath}
\begin{equation*}
    f(x) = 2 \times x
\end{equation*}
\end{linenomath}

The following is the XML document that encodes the model shown
above:

\sbmlexample{functiondef.xml}


%-----------------------------------------------------------------------------
\subsection{Example involving \emph{delay} functions}
\label{sec:delayeg}
%-----------------------------------------------------------------------------

The following is a simple model illustrating the use of $delay$ to
represent a gene that suppresses its own expression.  The model
can be expressed in a single rule:
\begin{linenomath}
\begin{equation*}
  \frac{d P}{d t} = \dfrac{ \dfrac{1}{1 + m (P_{delayed})^q} - P }{ \tau }
\end{equation*}
\end{linenomath}
where
\begin{linenomath}
\begin{equation*}
\begin{array}{rl}
P_{delayed} & \mbox{is } delay(P, \Delta_t) \mbox{ or P at } t - \Delta_t\\
P & \mbox{is protein concentration}\\
\tau & \mbox{is the response time}\\
m & \mbox{is a multiplier or equilibrium constant}\\
q & \mbox{is the Hill coefficient}\\
\end{array}
\end{equation*}
\end{linenomath}
and the species quantities are in concentration units.
The text of an SBML encoding of this model is given below:

\sbmlexample{delay.xml}


%-----------------------------------------------------------------------------
\subsection{Example involving events}
\label{sec:eventeg}
%-----------------------------------------------------------------------------

This section presents a simple model system that demonstrates the
use of events in SBML.  Consider a system with two genes, $k_1$
and $k_2$.  $k_1$ is initially on and $k_2$ is initially off.
When turned on, the two genes lead to the production of two
products, $P_1$ and $P_2$, respectively, at a fixed rate.  When
$P_1$ reaches a given concentration, $k_2$ switches off.  This
system can be represented mathematically as follows:
\begin{linenomath}
\begin{eqnarray*}
  \dfrac{d P_1}{d t} & = & k_1 - P_1\\[3pt]
  \dfrac{d P_2}{d t} & = & k_2 - P_2\\
  k_2 & = &
    \begin{cases}
      0 & \text{when $P_1 \leq \tau$},\\
      1 & \text{when $P_1 > \tau$}.
    \end{cases}
\end{eqnarray*}
\end{linenomath}

The initial values are:
\begin{linenomath}
\begin{equation*}
  k_1 = 1, \quad k_2 = 0, \quad \tau = 0.25, \quad P_1 = 0, \quad P_2 = 0
\end{equation*}
\end{linenomath}

The SBML Level 2 representation of this as follows:

\sbmlexample{events.xml}


%-----------------------------------------------------------------------------
\subsection{Example involving two-dimensional compartments}
\label{sec:two-dimensional-eg}
%-----------------------------------------------------------------------------

The following example is a model that uses a two-dimensional
compartment.  It is a fragment of a larger model of calcium
regulation across the plasma membrane of a cell.  The model
includes a calcium influx channel, \val{Ca\_channel}, and a
calcium-extruding PMCA pump, \val{Ca\_Pump}.  It also includes two
cytosolic proteins that buffer calcium via the
\val{CalciumCalbindin\_gt\_BoundCytosol} and
\val{CalciumBuffer\_gt\_BoundCytosol} reactions.  Finally, the
rate expressions in this model do not include explicit factors of
the compartment volumes; instead, the various rate constants are
assume to include any necessary corrections for volume.

\sbmlexample{twodimensional.xml}


%%-----------------------------------------------------------------------------
%\subsection{Example Using \SpeciesType Structures}
%\label{sec:speciesType-eg}
%%-----------------------------------------------------------------------------

%The following example is a model that uses \SpeciesType structures to indicate
%that 2 pools of biochemical entities (\emph{species}) located in different compartments
%contain the same type of entity (\emph{speciesType}).

%\begin{small}
%\tightspacing
%\begin{verbatim}
%<?xml version="1.0" encoding="UTF-8"?>
%<sbml xmlns="http://www.sbml.org/sbml/level2/version2" level="2" version="2">
%<model id="malate_aspartate_shuttle2">
%    <listOfCompartments>
%        <compartment id="Cytosol"/>
%        <compartment id="Mitochondrial_Matrix"/>
%    </listOfCompartments>
%    <listOfSpeciesTypes>
%        <speciesType id="Aspartate"/>
%    </listOfSpeciesTypes>
%    <listOfSpecies>
%        <species
%            id="Aspartate_in_Cytosol"
%            speciesType="Aspartate"
%            compartment="Cytosol"/>
%        <species
%            id="Aspartate_in_Mitochondrial_Matrix"
%            speciesType="Aspartate"
%            compartment="Mitochondrial_Matrix"/>
%    </listOfSpecies>
%</model>
%</sbml>
%\end{verbatim}
%\end{small}


