% -*- TeX-master: "main"; fill-column: 72 -*-

\section{Background and context}
\label{background}

The focus of this section is prior work on the topic of model composition in SBML.  We also explain how the current specification relates to that prior work.


\subsection{Prior work on model composition in SBML}
\label{prior-work}

The SBML community has discussed the need for model composition since SBML's very beginning, in 2000.  The approach to composition contained in the present document draws substantially on prior work.  Before we turn to a narrative of the history that led to this specification, we want to highlight a number of individuals for their inspirations and past work in the development of precursors to this package.  These individuals are listed in \tab{many-thanks}.

\begin{table}[hb]
  \centering
  \renewcommand{\arraystretch}{1.1}
  \rowcolors{2}{sbmlrowgray}{}
  \begin{edtable}{tabular}{lll}
    \toprule
    \textbf{Contributor}	& \textbf{Affiliation} & \textbf{City and Country}\\
    \midrule
    Frank Bergmann		& California Institute of Technology & Pasadena, CA, US\\
    Michael Blinov		& University of Connecticut Health Center & Farmington, CT, US\\
    Nicolas Le Nov\`{e}re	& EMBL European Bioinformatics Institute & Hinxton, Cambridge, UK\\
    Ranjit Randhawa		& Department of Computer Science, Virginia Tech. & Blacksburg, VA, US\\
    J\"{o}rg Stelling		& Max Planck Institute for Dynamics of Complex Technical Systems & Magdeburg, DE\\
    Jonathan Webb		& BBN Technologies & Cambridge, MA, US\\
    \bottomrule
  \end{edtable}
  \caption{Individuals who made significant contributions that influenced
    this Hierarchical Model Composition specification.}
  \label{many-thanks}
\end{table}

The first known written proposal for composition in SBML appeared in an internal discussion document titled \emph{Possible extensions to the   Systems Biology Markup Language}~\citep{finney:2000} principally authored by Andrew Finney (and, notably, written even before SBML Level~1 Version~1 was finalized in March of 2001).  The first of the four titular possible extensions in that document concerns ``submodels'': the main model in a file can contain a list of submodels, each of which are model definitions only, and a list of submodel instantiations, each of which are references to model definitions.  Finney's proposal also extends the syntax of SBML identifiers (the \primtype{SId} data type) to allow entity references using a dotted notation, in which \texttt{X.y} signifies element \texttt{y} of submodel instance \texttt{X}; the proposal also defines a form of linking model elements through ``substitutions''.  In addition, the proposal also introduces the concept of validation through what it called the ``expanded'' version of the model (now commonly referred to as the ``flattened'' form, meaning translation to a plain SBML format that does not use composition features): if the flat version of the model is valid, then the model as a whole must also be valid.

In June of 2001, at the Third Workshop on Software Platforms for Systems Biology, Martin Ginkel and J\"{o}rg Stelling presented their proposal titled \emph{XML Notation for Modularity}~\citep{ginkel:2001}, complete with a sample XML file.  Their efforts were in response to deficiencies or missing elements they believed existed in the proposal by Finney.  In their proposal, Ginkel and Stelling present a ``classic view'' of modularity, where models are packaged as black boxes with interfaces.  One of their design goals is to support the substitution of one module for another with the same defined interface, thereby supporting the simplification or elaboration of models as needed.  Their proposal emphasizes the reuse of models and with the possibility of developing libraries of models.

Martin Ginkel presented an expanded version of that proposal~\citep{ginkel:2002} at in the July 2002 Fifth Workshop on Software Platforms for Systems Biology, in the hope that it could be incorporated into the definition of SBML Level 2 that was being developed at the time.  This proposal clarified the need to separate model definitions from model instantiations, and, further, the need to designate one model per document as the ``main'' model.

In March of 2003, Jonathan Webb produced an independent proposal~\citep{webb:2003} and circulated it on the mailing list \link{http://sbml.org/Forums}{sbml-discuss@caltech.edu}.  This proposal included a unified, generic approach to making links and references to elements in submodels using XML XPath~\citep{xpath}.  Previous proposals used separate mechanisms for species, parameters, compartments, and reactions.  Webb also raised the issue of how to successfully resolve conflicting attributes of linked elements, debated whether formal interfaces were necessary or even preferable to directly accessing model elements, discussed type-checking for linkages, and discussed issues with unit incompatibilities.  Around this time, Martin Ginkel formed the Model Composition Special Interest Group~\citep{comp-sig}, a group that eventually reached 18 members (including Webb).

Model composition did not make it into SBML Level 2 when that specification was released in June of 2003, because the changes between SBML Level~1 and Level~2 were already substantial enough that software developers at the time expressed a desire to delay the introduction of composition to a later revision of SBML.  Andrew Finney (now the co-chair of the Model Composition SIG) presented yet another proposal~\citep{finney:2003} in May of 2003, even before SBML Level~2 Version~1 was finalized, that aimed to add model composition to SBML Level~3.  With only two years having passed between SBML Level~1 and Level~2, the feeling at the time was that Level~3 was likely to be released in 2005 or 2006, and the model composition proposal would be ready when it was.  However, Level~2 ended up occupying the SBML community longer than expected, with four versions of Level~2 produced to adjust features in response to user feedback and developers' experiences.

In the interim, the desire to develop model composition features for SBML continued unabated.  Finney revised his 2003 proposal in October 2003~\citep{finney:2003b}; this new version represented an attempt to synthesize the earlier proposals by Ginkel and Webb, supplemented with his own original submodel ideas, and was envisioned to exist in parallel with another proposal by Finney, for arrays and sets of SBML elements (including submodels)~\citep{finney:2003c}.  Finney attempted to resolve the differences in the two basic philosophies (essentially, black-box versus white-box encapsulation) by introducing optional ``ports'' as interfaces between a submodel and its containing model, as well as including an XPath-based method to allow referencing model entities. The intention was that a modeler who wanted to follow the classic modularity (black-box) approach could do so, but other modelers could still use models in ways not envisioned by the original modeler simply by accessing a model's elements directly via XPath-based references.  In both schemes, elements in the submodels were replaced by corresponding elements of the containing model.  Finney's proposal also provided a direct link facility that allows a containing model to refer directly to submodel elements without providing placeholder elements in the containing model.  For example, a containing model could have a reaction that converts a species in one submodel to a species in a different submodel, and in the direct-link approach, it would only need to define the reaction, with the reactant and product being expressed as links directly to the species defined in the submodels. 

After Finney's last effort, activities in the SBML community focused on updates to SBML Level~2, and since model composition was slated for Level~3, not much progress was made for several years, apart from Finney including a summary of his 2003 proposal and of some of the unresolved issues in a poster~\citep{finney:2004} at the 2004 Intelligent Systems for Molecular Biology (ISMB) conference held in Glasgow.

Finally, in June, 2007, unplanned discussions at the Fifth SBML Hackathon~\citep{sbml5} prompted the convening of a workshop to revitalize the model composition package, and in September of 2007, the SBML Composition Workshop~\citep{comp2007} was held at the University of Connecticut Health Center, hosted by the Virtual Cell group and organized by Ion Moraru and Michael Blinov.  The event produced several artifacts:

\begin{enumerate}

\item Martin Ginkel provided a list of goals for model composition~\citep{ginkel:2007}, including use cases, and summarized many of the issues described above, including the notion of definition versus instantiation, linking, referencing elements that lack SBML identifiers, and the creation of optional interfaces.  The list of goals also mentioned the need of allowing parameterization of instances (i.e., setting new numerical values that override the defaults), and the need to be able to ``delete'' or elide elements out of submodels.  (He also provided a summary of ProMoT's model composition approach and a summary of other approaches.)

\item Andrew Finney created a list of issues and comments, recorded on the meeting's wiki page~\citep{finney:2007}; the list included some old issues as well as some new ones:

  \begin{itemize}

  \item There should perhaps be a flag for ports to indicate whether a given port must be overloaded.

  \item There should be support for N-to-M links, when a set of elements in one model are replaced as a group, conceptually, with one or more elements from a different model. 

  \item The proposal should be generic enough to accommodate future updates and other Level~3 packages.

  \end{itemize}
  
\item Wolfram Liebermeister presented his group's experience with SBMLMerge~\citep{liebermeister:2007}, dealing with the pragmatics of merging multiple models.  He also noted that the annotations in a composed model need to be considered, particularly since they can be crucial to successfully merging models in the first place.

\item On behalf of Ranjit Randhawa, Cliff Shaffer summarized Ranjit's work in the JigCell group on model fusion, aggregation, and composition~\citep{randhawa:2007}.  Highlights included the following:

  \begin{itemize}

  \item A description of different methods which all need some form of model composition, along with the realization that model fusion and model composition, though philosophically different, entail exactly the same processes and require the same information.

  \item A software application (the JigCell Composition Wizard) that can perform conversion between types.  The application can, for example, promote a parameter to a species, a concept which had been assumed to be impossible and undesirable in previous proposals.  

  \item The discovery that merging of SBML models should be done in the order Compartments $\rightarrow$ Species  $\rightarrow$ Function Definitions  $\rightarrow$ Rules  $\rightarrow$ Events $\rightarrow$ Units  $\rightarrow$ Reactions  $\rightarrow$ Parameters.  If done in this order, potential conflicts are resolved incrementally along the way.

  \end{itemize}

\item Nicolas Le~Nov\`{e}re created a proposal for SBML modularity in Core~\citep{lenov:2007}.  This is actually unrelated to the efforts described above; it is an attempt to modularize a ``normal'' SBML model in the sense of divvying up the information into modules or blocks stored in separate files, rather than composing a model from different chunks.  It was agreed at the workshop that this is a completely separate idea, and while it has merits, should be handled separately.

\item The group produced an ``Issues to Address'' document~\citep{various:2007}, with several conclusions:

  \begin{itemize}

  \item It should be possible to ``flatten'' a composed model to produce a valid SBML Level 3 Core model, and all questions of validity can then be simply applied to the flattened model.  If the Core-only version is valid, the composed model is valid.

  \item The model composition proposal should cover both designed-ahead-of-time as well as ad-hoc composition. (The latter refers to composing models out of components that were not originally developed with the use of ports or the expectation of being incorporated into other models.)

  \item The approach probably needs a mechanism for deleting SBML model elements.  The deletion syntax should be explicit, instead of being implied by (e.g.)\ using a generic replacement construct and omitting the target of the replacement.

  \item It should be possible to link any part of a model, not just (e.g.)\ compartments, species and parameters.

  \item The approach should support item ``object overloading''~\citep{various:2007b} and be generally applicable to all SBML objects.  However, contrary to what is provided in the JigCell Composition Wizard, changing SBML component types is not supported in object overloading.

  \item A proposition made during the workshop is that elements in the outer model always override elements in the submodels, and perhaps that sibling linking be disallowed.  This idea was hotly debated.

  \item Interfaces (ports) are considered helpful, but optional.  They do not need to be directional as in the electrical engineering ``input'' and ``output'' sense; the outer element always overrides the inner element, but apart from that, biology does not tend to work in the directional way that electrical components do.

  \item The ability to refer to or import external files may need a mechanism to allow an application to check whether what is being imported is the same as it was when the modeler created the model.  The mechanism offered in this context was the use of MD5 hashes.

  \item A model composition approach should probably only allow whole-model imports, not importing of individual SBML elements such as species or reactions.  Model components are invariably defined within a larger context, and attempting to pull a single piece out of a model is unlikely to be safe or desirable.

  \item The approach must provide a means for handling unit conversions, so that the units of entities defined in a submodel can be made congruent with those of entities that refer to them in the enclosing model.

  \end{itemize}

\end{enumerate}

During the workshop, the attendees worked on a draft proposal.  Stefan Hoops acted as principal editor.  The proposal for the SBML package (which was renamed \emph{Hierarchical Model Composition}~\citep{hoops:2007}), was issued one day after the end of the workshop.  It represented an attempt to summarize the workshop as a whole, and provide a coherent whole, suitable as a Level~3 package.  It provided a brief overview of the history and goals of the proposal, as well as several UML diagrams of the proposed data structures.  Hoops presented~\citep{hoops:2008} the proposal in August, 2008, at the 13th SBML Forum, and again at the 7th SBML Hackathon in March of 2009 as well as the 14th SBML Forum in September of 2009, in a continuing effort to raise interest.

Roughly concurrently, Herbert Sauro, one of the original developers of SBML, received a grant to develop a modular human-readable model definition language, and hired Lucian Smith in November of 2007 to work on the project.  Sauro and Frank Bergmann, then a graduate student with Herbert, had previously written a proposal~\citep{bergmann:2006} for a human-readable language that provided composition features, and this was the design document Smith initially used to create a software system that was eventually called \link{http://antimony.sourceforge.net/}{Antimony}. Through a few iterations, the design eventually settled on was very similar in concept (largely by coincidence) to that developed by the group at the 2007 Connecticut workshop: namely, with model definitions placed separately from their instantiations in other models, and with the ability to link (or ``synchronize'', in Antimony terminology) elements of models with each other.  Because Antimony was designed to be ``quick and dirty'', it allowed type conversions much like the JigCell Composition Wizard, whereby a parameter could become a species, compartment, or even reaction.  Synchronized elements could end up with aspects of both parent elements in their final definitions: if one element defined a starting condition and the other how it changed in time, the final element would have both.  If both elements defined the same aspect (like a starting condition), the one designated the ``default'' would be used in the final version.  Smith developed methods to import other Antimony files and even SBML models, which could then be used as submodels of other models and exported as flattened SBML.

At the SBML-BioModels.net Hackathon in 2010, in response to popular demand from people who attended the workshop, Smith put together a short presentation~\citep{smith:2010} about model composition and some of the limitations he found with the 2007 proposal.  He proposed separating the replacement concept (where old references to replaced values are still valid) from the deletion concept (where old references to replaced values are no longer valid).  Smith wrote a summary of that discussion, added some more of thoughts, and posted it to the \link{http://sbml.org/Forums}{sbml-discuss@caltech.edu} mailing list~\citep{smith:2010b}.  In this posting, he proposed and/or reported several possible modifications to the Hoops et al.\ 2007 proposal, including the following:

\begin{itemize}\setlength{\parskip}{0ex}

\item Separation of \emph{replacement} from \emph{deletion}.

\item Separation of model definition from instantiation.

\item Elimination of ports, and the use of annotations instead.

\item Annotation for identifying N-to-M replacements, instead of giving them their own construct.

\end{itemize}

The message to \link{http://sbml.org/Forums}{sbml-discuss@caltech.edu} was met with limited discussion.  However, it turns out that several of the issues raised by Smith were brought up at the 2007 meeting, and had simply been missed in the generation of the (incomplete) proposal after the workshop.  The meeting attendees had, for example, originally preferred to differentiate deletions from replacements more strongly than by simply having an empty list of replacements, but omitted this feature because no better method could be found.  Similarly, the separation of definitions from instantiations had been in every proposal up until 2007, and was mentioned in the notes for that meeting.  The decision to merge the two was a last-minute design decision brought about when the group noted that if the XInclude~\citep{xinclude} construct was used, the separation was not strictly necessary from a technical standpoint.

Smith joined the SBML Team in September of 2010, and was tasked with going through the old proposals and synthesizing from them a new version that would work with the final incarnation of SBML Level 3.  He presented that work at COMBINE in October 2010~\citep{smith:2010c}, and further discussed it on the mailing list \link{http://sbml.org/Forums}{sbml-discuss@caltech.edu}.  At HARMONY in April of 2011, consensus was reached on a way forward for resolving the remaining controversies surrounding the specification, resulting in the first draft of this document.

\subsection{Genesis of \changed{Version 1} of this specification}

\changed{Version 1 of this} specification for Hierarchical Model Composition was an attempt to blend features of previous efforts into a concrete, Level~3-compatible syntax.  The specification was written from scratch, but drew strongly on the Hoops 2007 and Finney 2003 proposals, as well as, to some degree, every one of the sources mentioned above. Some practical decisions were new to this proposal, sometimes due to additional design constraints resulting from the final incarnation of SBML Level~3, but all of them drew from a wealth of history and experimentation by many different people over the previous decade.  Where this proposal differs from the historical consensus, the reasoning is explained, but for the most part, the proposal follows the road most traveled, and focuses on being clear, simple, only as complex as necessary, and applicable to the largest number of situations.

The first draft of this specification was discussed at COMBINE 2011, during which time several concerns were raised by Chris Myers and Frank Bergmann.  These included the following: (1) the semantics of a Boolean attribute, \token{identical}, on \ReplacedElement; (2) the high complexity of conversion factors as they were defined at the time; (3) the inelegance of recursive chains of \SBaseRef objects; and (4) the need for MD5 checksums.  Those discussions resulted in the following conclusions:

\begin{enumerate}

\item The Boolean attribute \token{identical} could be removed from the definition of \ReplacedElement, and in its place a new construct added, \ReplacedBy.

\item The conversion factor scheme in the original first draft was indeed dauntingly complex.

\item There was general agrement that recursive chains of \SBaseRef objects is an inelegant approach, but a workable and relatively simple one.  \emph{Some} mechanism of that sort is necessary if we accept that one of the use cases for Hierarchical Model Composition involves externally-referenced models that a modeler does not own and cannot change.

\item For the same use case reasons, MD5 checksums are sufficiently important to leave in.  If a referenced external model ever changes without a modeler's knowledge, checksums can allow a software system to detect and report the situation.

\end{enumerate}

To address issue (2), Smith set out to develop a revised and much simplified version that reduced the number of conversions factors considerably.  This scheme was incorporated into a revised version of a revised draft specification issued in 2012.  Smith then worked on implementating this draft specification in libSBML and integrating it into \link{http://antimony.sourceforge.net/}{Antimony}, and Chris Myers worked on integrating it into \link{http://www.async.ece.utah.edu/iBioSim/}{iBioSim}.  As a result of these experiences, additional issues became apparent and were discussed during HARMONY 2012.  Those dicussions produced the following conclusions and changes:

\begin{enumerate}

\item The use of ports is not required and is a relatively foreign concept to many biologists; thus, their use is more of a best-practices matter, and a requirement to use ports cannot be imposed.

\item The previous formulation of \ReplacedElement and \ReplacedBy, which allowed replacement of any objects without restriction, led to complex and awkward software interfaces, and made it easy to create invalid models.  On the other hand, restricting replacements to \emph{only} involve the same classes of objects (``like-with-like'' replacements) was deemed too restrictive.  The compromise solution was to only allow parameters to replace other classes of SBML objects that had a mathematical meaning.  In all other cases, objects can only be replaced (or act as replacements for) objects of the same class or subclasses of that class.

\end{enumerate}



\subsection{Design goals for the Hierarchical Model Composition package}
\label{sec:design-goals}

The following are the basic design goals followed in this package:

\begin{itemize}

\item \emph{Allow modelers to build models by aggregation, composition, or modularity}.  These methods are so similar to one another, and the process of creating an SBML Level 3 package is so involved, that we believe it is not advantageous to create one SBML package for aggregation and composition, and a separate package for modularity. Users of the hierarchical model composition package should be able to use and create models in the style that is best suited for their individual tasks, using any of these mechanisms, and to exchange and reuse models from other groups simply and straightforwardly.

\item \emph{Interoperate cleanly with other packages}. The rules of composition should be such that they could apply to any SBML element, even unanticipated elements not defined in SBML Level 3 Core and introduced by some future Level 3 package.

\item \emph{Allow models produced with these constructs to be valid SBML if the constructs are ignored}.  As proposed by \citet*{lenov:2003} and affirmed by the SBML Editors~\citep{editors:2010}, whenever possible, ignoring elements defined in a Level 3 package namespace should result in syntactically-correct SBML models that can still be interpreted to some degree, even if it cannot produce the intended simulation results of the full (i.e., interpreting the package constructs) model.  For example, inspection and visualization of the Core model should still be possible.

\item \emph{Ignore verbosity of models}. We assume that software will deal with the ``nuts and bolts'' of reading and writing SBML.  If there are two approaches to designing a mechanism for this hierarchical composition package, where one approach is clear but verbose and the other approach is concise but complex or unobvious, we prefer the clear and verbose approach.  We assume that software tools can abstract away the verbosity for the user.  (However, tempering this goal is the next point.)

\item \emph{Avoid over-complicating the specification}. Apart from the base constructs defined by this specification, any new element or attribute introduced should have a clear use case that cannot be achieved in any other way.

\item \emph{Allow modular access to files outside the modeler's control}.  In order to encourage direct referencing of models (e.g., to models hosted online on sites such as BioModels Database (\url{http://biomodels.net/database}), whenever possible, we will require referenced submodels only to be in SBML Level~3 format (with or without other package information), and not require that they include constructs from this specification.

\item \emph{Incorporate most, if not all, of the desirable features of past proposals}. The names may change, but the aims of past efforts at SBML model composition should still be achievable with the present specification.

\end{itemize}

\begin{blockChanged}
Release~1 of this specification was approved by the SBML editors in November of 2012, with minor updates and corrections following in Release~2 in February of 2013, and Release~3 in November of 2013.

\subsection{Genesis of \changed{Version 2} of this specification}

The main driving force behind Version~2 of this specification was the development and release of SBML Level~3 Version~2 Core.  The main change there that necessitated an update for the Hierarchical Model Composition specification was the introduction of the \token{id} and \token{name} attributes on \SBase, the abstract base class that most elements in SBML inherit from.  To be compliant with that change, all \token{comp:id} and \token{comp:name} attributes defined here were dropped, in favor of the new attributes in the core namespace.

The second design change that was incorported into this specification was the relaxation of the restrictions on external model definitions:  now, any SBML document of any level or version may be referenced by an \ExternalModelDefinition, as long as that model can be unambiguously and losslessly translated to SBML Level~3 Version~2.  While there are many edge cases that could indeed be ambiguous and/or lossy (such as when one transates an SBML Level~2 \StoichiometryMath to an SBML Level~3 \Rule), those situations are rare and not likely to be encountered by modelers on a regular basis.  When they do come up, it is possible to simply translate the model by hand first, and then reference that translated model with an \ExternalModelDefinition.  For this reason, the benefit of being able to reference older models in a hierarchical context was deemed to outweigh the costs associated with ambiguities in translation.

\end{blockChanged}
