\documentclass{jib}
\newlength{\platz}
\setlength{\platz}{15pt}
\RequirePackage{listings}

\usepackage{changepage} %test, TODO remove

\lstset{%
  basicstyle=\ttfamily,
  fontadjust,
  flexiblecolumns=true,
  frame=L,
  xleftmargin=15pt,
  framesep=5pt,
  emphstyle=\rmfamily\itshape}

\usepackage{pdfpages}  

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% JIB Header/Footer
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\jibvolume{XX} % insert volume
%\jibissue{X}   % insert issue
%\jibpages{XXX} % insert article ID
%\jibyear{XXXX} % insert year
%\makeHeaderFooter{} % leave as is
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Title Page
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{jibtitlepage}

\jibtitle{SBML Level 3 package: Multistate, Multicomponent and Multicompartment Species, Version 1, Release 2}


%We did not provide author(s) nor author footnote(s), please complete as applicable.
% Please make sure to use unique footnote characters for each author

\jibauthor{ 
  Fengkai Zhang\iref{niaid},
  Lucian P. Smith\iref{uw},
  Michael L. Blinov\iref{uconn},
  James Faeder\iref{upitt},
  William S. Hlavacek\iref{lanl},
  Jose Juan Tapia\iref{upitt},
  Sarah M. Keating\iref{ebi},
  Nicolas Rodriguez\iref{babraham},
  Andreas Dr{\"a}ger\iref{tuebingen},
  Leonard A. Harris\iref{vusm},
  Andrew Finney\iref{aul},
  Bin Hu\iref{lanl},
  Michael Hucka\iref{caltech},
  Martin Meier-Schellersheim\iref{niaid}
}


\addjibinstitution{niaid}{NIAID/NIH, US}
\addjibinstitution{uw}{University of Washington, US}
\addjibinstitution{uconn}{UConn Health, US}
\addjibinstitution{upitt}{University of Pittsburgh, US}
\addjibinstitution{lanl}{Los Alamos National Laboratory, US}
\addjibinstitution{ebi}{European Bioinformatics Institute, GB}
\addjibinstitution{babraham}{Babraham Institute, GB}
\addjibinstitution{tuebingen}{University of T{\"u}bingen, DE}
\addjibinstitution{vusm}{Vanderbilt University School of Medicine, US}
\addjibinstitution{aul}{Anasys UK Ltd, GB}
\addjibinstitution{caltech}{California Institute of Technology, US}



\end{jibtitlepage}


% \begin{abstract}
% 
% This document is a short description of the \LaTeX\ \emph{document class} \emph{jib} and its use. It shall be used to submit \LaTeX\ articles to the \emph{Journal of Integrative Bioinformatics} and at the same time may be used to check whether your \LaTeX\ installation can compile output files according to the guidelines of the journal.
% 
% \end{abstract}

% adjusts the width of the abstract, please do not change!
\begin{adjustwidth}{}{1cm}                 

\abstract{
Rule-based modeling is an approach that permits constructing reaction
networks based on the specification of rules for molecular
interactions and transformations. These rules can encompass details
such as the interacting sub-molecular 
domains (components) and the
states such as phosphorylation and binding  
status of the involved
components. Conceptually, fine-grained spatial information such as the locations of
the molecular components relative to a membrane (e.g. whether a
modeled molecular domain is embedded into the inner leaflet of the 
cellular plasma membrane) can also be provided. Through ``wildcards''
representing component states, entire families of molecule complexes
sharing certain properties can be specified as patterns. This can
significantly simplify the definition of models involving species with
multiple components, multiple states, and multiple compartments. 
The SBML Level~3 \emph{Multi} Package (Multistate, Multicomponent and
Multicompartment Species Package for SBML Level~3) Version~1 extends the SBML Level~3 Version 1
\emph{core} with the ``type'' concept in the Species and Compartment classes. Therefore, 
reaction rules may contain species that can be patterns
and exist in multiple locations. 
Multiple software tools such as \emph{Simmune} and \emph{BioNetGen} support the SBML
Level~3 \emph{Multi} package that thus also becomes a medium for exchanging rule-based models. This document provides the specification for Release 2 of Version 1 of the SBML Level~3 \emph{Multi} package. No design changes have been made to the description of models between Release 1 and Release 2; changes are restricted to the correction of errata and the addition of clarifications. All the completed example models in this specification have been validated by the validation rules of SBML Level 3 Version 1 core and SBML Level 3 Multi package Version 1 using the libSBML library.
}

\end{adjustwidth} % please do not change

\end{document}
