% -*- TeX-master: "main"; fill-column: 72 -*-

\section{Validation of SBML documents}
\label{apdx-validation}

\subsection{Validation and consistency rules}
\label{validation-rules}

This section summarizes all the conditions that must (or in some cases,
at least \emph{should}) be true of an SBML Level~3 Version~2 model that
uses the \DistributionsPackage. We use the same conventions as are used
in the SBML Level~3 Version~2 Core specification document. In
particular, there are different degrees of rule strictness. Formally,
the differences are expressed in the statement of a rule: either a rule
states that a condition \emph{must} be true, or a rule states that it
\emph{should} be true. Rules of the former kind are strict SBML
validation rules---a model encoded in SBML must conform to all of them
in order to be considered valid. Rules of the latter kind are
consistency rules. To help highlight these differences, we use the
following three symbols next to the rule numbers:

\begin{description}

\item[\hspace*{6.5pt}\vSymbol\vsp] A \vSymbolName indicates a
\emph{requirement} for SBML conformance. If a model does not follow this
rule, it does not conform to the \DistributionsPackage specification.
(Mnemonic intention behind the choice of symbol: ``This must be
checked.'')

\item[\hspace*{6.5pt}\cSymbol\csp] A \cSymbolName indicates a
\emph{recommendation} for model consistency. If a model does not follow
this rule, it is not considered strictly invalid as far as the
\DistributionsPackage specification is concerned; however, it indicates
that the model contains a physical or conceptual inconsistency.
(Mnemonic intention behind the choice of symbol: ``This is a cause for
warning.'')

\item[\hspace*{6.5pt}\mSymbol\msp] A \mSymbolName indicates a strong
recommendation for good modeling practice. This rule is not strictly a
matter of SBML encoding, but the recommendation comes from logical
reasoning. As in the previous case, if a model does not follow this
rule, it is not strictly considered an invalid SBML encoding. (Mnemonic
intention behind the choice of symbol: ``You're a star if you heed
this.'')

\end{description}

The validation rules listed in the following subsections are all stated
or implied in the rest of this specification document. They are
enumerated here for convenience. Unless explicitly stated, all
validation rules concern objects and attributes specifically defined in
the \DistributionsPackage package.

For \notice convenience and brevity, we use the shorthand
``\token{distrib:\-x}'' to stand for an attribute or element name
\token{x} in the namespace for the \DistributionsPackage package, using
the namespace prefix \token{distrib}. In reality, the prefix string may
be different from the literal ``\token{distrib}'' used here (and indeed,
it can be any valid XML namespace prefix that the modeler or software
chooses). We use ``\token{distrib:\-x}'' because it is shorter than to
write a full explanation everywhere we refer to an attribute or element
in the \DistributionsPackage namespace.

\changed{Attributes from this package are listed in these rules as having the \val{distrib:} prefix, but as is convention for SBML packages, this prefix is optional.}

\subsubsection*{General rules about this package}

\validRule{distrib-10101}{To conform to the \DistributionsPackage
specification for SBML Level~3 Version~2, an SBML document must declare
\uri{http://www.sbml.org/sbml/level3/version2/distrib/version1} as the
XMLNamespace to use for elements of this package. (Reference: SBML
Level~3 Package specification for Distributions, Version~1
\sec{xml-namespace}.)}

\validRule{distrib-10102}{Wherever they appear in an SBML document,
elements and attributes from the \DistributionsPackage must use the
\uri{http://www.sbml.org/sbml/level3/version2/distrib/version1}
namespace, declaring so either explicitly or implicitly. (Reference:
SBML Level~3 Package specification for Distributions, Version~1
\sec{xml-namespace}.)}

\subsubsection*{General rules about identifiers}

\validRule{distrib-10301}{(Extends validation rule \#10301 in the
\sbmlthreecore specification. TO DO list scope of ids) (Reference: SBML
Level~3 Version~1 Core, Section~3.1.7.)}

\validRule{distrib-10302}{The value of a \token{distrib:\-id} must
conform to the syntax of the \class{SBML} data type \primtype{SId}
(Reference: SBML Level~3 Version~1 Core, Section~3.1.7.)}

\TODO{ANY LIST OF ELEMENTS THAT HAVE ATTRIBUTES}

\subsubsection*{Rules for the extended \class{SBML} class}

\validRule{distrib-20101}{In all SBML documents using the
\DistributionsPackage, the \class{SBML} object must have the
\token{distrib:\-required} attribute. (Reference: SBML Level~3 Version~1
Core, Section~4.1.2.)}

\validRule{distrib-20102}{The value of attribute
\token{distrib:\-required} on the \class{SBML} object must be of data
type \primtype{boolean}. (Reference: SBML Level~3 Version~1 Core,
Section~4.1.2.)}

\validRule{distrib-20103}{The value of attribute
\token{distrib:\-required} on the \class{SBML} object must be set to
\val{true}. (Reference: SBML Level~3 Package specification for
Distributions, Version~1 \sec{xml-namespace}.)}

%\subsubsection*{Rules for extended \class{FunctionDefinition} object}

%\validRule{distrib-20201}{A \FunctionDefinition object may contain one
%and only one instance of the \DrawFromDistribution element. No other
%elements from the SBML Level 3 Distributions namespaces are permitted on
%a \FunctionDefinition object. (Reference: SBML Level~3 Package
%specification for Distributions, Version~1,
%\sec{extended-functiondefinition-class}.)}


\subsubsection*{Rules for extended \class{SBase} object}

\validRule{distrib-20301}{A \SBase object may have the optional
attributes \token{distrib:\-id} and \token{distrib:\-name}. No other
attributes from the SBML Level 3 Distributions namespaces are permitted
on a \SBase object. (Reference: SBML Level~3 Package specification for
Distributions, Version~1, \sec{extended-sbase-class}.)}

\validRule{distrib-20302}{A \SBase object may contain one and only one
instance of the \Uncertainty element. No other elements from the SBML
Level 3 Distributions namespaces are permitted on a \SBase object.
(Reference: SBML Level~3 Package specification for Distributions,
Version~1, \sec{extended-sbase-class}.)}

\validRule{distrib-20303}{The attribute \token{distrib:\-name} on a
\SBase must have a value of data type \token{string}. (Reference: SBML
Level~3 Package specification for Distributions, Version~1,
\sec{extended-sbase-class}.)}


%\subsubsection*{Rules for \class{DrawFromDistribution} object}

%\validRule{distrib-20401}{A \DrawFromDistribution object may have the
%optional SBML Level~3 Core attributes \token{metaid} and
%\token{sboTerm}. No other attributes from the SBML Level 3 Core
%namespaces are permitted on a \DrawFromDistribution. (Reference: SBML
%Level~3 Version~1 Core, Section~3.2.)}

%\validRule{distrib-20402}{A \DrawFromDistribution object may have the
%optional SBML Level~3 Core subobjects for notes and annotations. No
%other elements from the SBML Level 3 Core namespaces are permitted on a
%\DrawFromDistribution. (Reference: SBML Level~3 Version~1 Core,
%Section~3.2.)}

%\validRule{distrib-20403}{A \DrawFromDistribution object may contain one
%and only one instance of each of the \ListOfDistribInputs and
%Distribution elements. No other elements from the SBML Level 3
%Distributions namespaces are permitted on a \DrawFromDistribution
%object. (Reference: SBML Level~3 Package specification for
%Distributions, Version~1, \sec{drawfromdistribution-class}.)}

%\validRule{distrib-20404}{Apart from the general notes and annotations
%subobjects permitted on all SBML objects, a \ListOfDistribInputs
%container object may only contain \DistribInput objects. (Reference:
%SBML Level~3 Package specification for Distributions, Version~1,
%\sec{listofdistribinputs-class}.)}

%\validRule{distrib-20405}{A \ListOfDistribInputs object may have the
%optional SBML Level~3 Core attributes \token{metaid} and
%\token{sboTerm}. No other attributes from the SBML Level 3 Core
%namespaces are permitted on a \ListOfDistribInputs object. (Reference:
%SBML Level~3 Package specification for Distributions, Version~1,
%\sec{listofdistribinputs-class}.)}


%\subsubsection*{Rules for \class{DistribInput} object}

%\validRule{distrib-20501}{A \DistribInput object may have the optional
%SBML Level~3 Core attributes \token{metaid} and \token{sboTerm}. No
%other attributes from the SBML Level 3 Core namespaces are permitted on
%a \DistribInput. (Reference: SBML Level~3 Version~1 Core, Section~3.2.)}

%\validRule{distrib-20502}{A \DistribInput object may have the optional
%SBML Level~3 Core subobjects for notes and annotations. No other
%elements from the SBML Level 3 Core namespaces are permitted on a
%\DistribInput. (Reference: SBML Level~3 Version~1 Core, Section~3.2.)}

%\validRule{distrib-20503}{A \DistribInput object may have the optional
%attributes \token{distrib:\-id}, \token{distrib:\-name} and
%\token{distrib:\-index}. No other attributes from the SBML Level 3
%Distributions namespaces are permitted on a \DistribInput object.
%(Reference: SBML Level~3 Package specification for Distributions,
%Version~1, \sec{distribinput-class}.)}

%\validRule{distrib-20504}{The attribute \token{distrib:\-name} on a
%\DistribInput must have a value of data type \token{string}. (Reference:
%SBML Level~3 Package specification for Distributions, Version~1,
%\sec{distribinput-class}.)}

%\validRule{distrib-20505}{The attribute \token{distrib:\-index} on a
%\DistribInput must have a value of data type \token{integer}.
%(Reference: SBML Level~3 Package specification for Distributions,
%Version~1, \sec{distribinput-class}.)}


\subsubsection*{Rules for \class{Distribution} object}

\validRule{distrib-20601}{A \Distribution object may have the optional
SBML Level~3 Core attributes \token{metaid} and \token{sboTerm}. No
other attributes from the SBML Level 3 Core namespaces are permitted on
a \Distribution. (Reference: SBML Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-20602}{A \Distribution object may have the optional
SBML Level~3 Core subobjects for notes and annotations. No other
elements from the SBML Level 3 Core namespaces are permitted on a
\Distribution. (Reference: SBML Level~3 Version~1 Core, Section~3.2.)}


\subsubsection*{Rules for \class{UnivariateDistribution} object}

\validRule{distrib-20701}{An \UnivariateDistribution object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on an \UnivariateDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-20702}{An \UnivariateDistribution object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on an
\UnivariateDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}


\subsubsection*{Rules for \class{MultivariateDistribution} object}

\validRule{distrib-20801}{A \MultivariateDistribution object may have
the optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \MultivariateDistribution. (Reference:
SBML Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-20802}{A \MultivariateDistribution object may have
the optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on a
\MultivariateDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}


\subsubsection*{Rules for \class{ContinuousUnivariateDistribution}
object}

\validRule{distrib-20901}{A \ContinuousUnivariateDistribution object may
have the optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \ContinuousUnivariateDistribution.
(Reference: SBML Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-20902}{A \ContinuousUnivariateDistribution object may
have the optional SBML Level~3 Core subobjects for notes and
annotations. No other elements from the SBML Level 3 Core namespaces are
permitted on a \ContinuousUnivariateDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-20903}{A \ContinuousUnivariateDistribution object may
contain one and only one instance of each of the UncertBound and
UncertBound elements. No other elements from the SBML Level 3
Distributions namespaces are permitted on a
\ContinuousUnivariateDistribution object. (Reference: SBML Level~3
Package specification for Distributions, Version~1,
\sec{continuousunivariatedistribution-class}.)}


\subsubsection*{Rules for \class{DiscreteUnivariateDistribution} object}

\validRule{distrib-21001}{A \DiscreteUnivariateDistribution object may
have the optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \DiscreteUnivariateDistribution.
(Reference: SBML Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-21002}{A \DiscreteUnivariateDistribution object may
have the optional SBML Level~3 Core subobjects for notes and
annotations. No other elements from the SBML Level 3 Core namespaces are
permitted on a \DiscreteUnivariateDistribution. (Reference: SBML Level~3
Version~1 Core, Section~3.2.)}

\validRule{distrib-21003}{A \DiscreteUnivariateDistribution object may
contain one and only one instance of each of the UncertBound and
UncertBound elements. No other elements from the SBML Level 3
Distributions namespaces are permitted on a
\DiscreteUnivariateDistribution object. (Reference: SBML Level~3 Package
specification for Distributions, Version~1,
\sec{discreteunivariatedistribution-class}.)}


\subsubsection*{Rules for \class{CategoricalUnivariateDistribution}
object}

\validRule{distrib-21101}{A \CategoricalUnivariateDistribution object
may have the optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \CategoricalUnivariateDistribution.
(Reference: SBML Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-21102}{A \CategoricalUnivariateDistribution object
may have the optional SBML Level~3 Core subobjects for notes and
annotations. No other elements from the SBML Level 3 Core namespaces are
permitted on a \CategoricalUnivariateDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}


\subsubsection*{Rules for \class{UncertValue} object}

\validRule{distrib-21201}{An \UncertValue object may have the optional
SBML Level~3 Core attributes \token{metaid} and \token{sboTerm}. No
other attributes from the SBML Level 3 Core namespaces are permitted on
an \UncertValue. (Reference: SBML Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-21202}{An \UncertValue object may have the optional
SBML Level~3 Core subobjects for notes and annotations. No other
elements from the SBML Level 3 Core namespaces are permitted on an
\UncertValue. (Reference: SBML Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-21203}{An \UncertValue object may have the optional
attributes \token{distrib:\-value}, \token{distrib:\-var} and
\token{distrib:\-units}. No other attributes from the SBML Level 3
Distributions namespaces are permitted on an \UncertValue object.
(Reference: SBML Level~3 Package specification for Distributions,
Version~1, \sec{uncertvalue-class}.)}

\validRule{distrib-21204}{The attribute \token{distrib:\-value} on an
\UncertValue must have a value of data type \token{double}. (Reference:
SBML Level~3 Package specification for Distributions, Version~1,
\sec{uncertvalue-class}.)}

\validRule{distrib-21205}{The value of the attribute
\token{distrib:\-var} of an \UncertValue object must be the identifier
of an existing \SBase object defined in the enclosing \Model object.
(Reference: SBML Level~3 Package specification for Distributions,
Version~1, \sec{uncertvalue-class}.)}

\validRule{distrib-21206}{The value of the attribute
\token{distrib:\-units} on an \UncertValue must have a taken from the
following: the identifier of a \UnitDefinition object in the enclosing
\Model, or one of the base units in SBML. (Reference: SBML Level~3
Package specification for Distributions, Version~1,
\sec{uncertvalue-class}.)}


\subsubsection*{Rules for \class{UncertBound} object}

\validRule{distrib-21301}{An \UncertBound object may have the optional
SBML Level~3 Core attributes \token{metaid} and \token{sboTerm}. No
other attributes from the SBML Level 3 Core namespaces are permitted on
an \UncertBound. (Reference: SBML Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-21302}{An \UncertBound object may have the optional
SBML Level~3 Core subobjects for notes and annotations. No other
elements from the SBML Level 3 Core namespaces are permitted on an
\UncertBound. (Reference: SBML Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-21303}{An \UncertBound object must have the required
attribute \token{distrib:\-inclusive}. No other attributes from the SBML
Level 3 Distributions namespaces are permitted on an \UncertBound
object. (Reference: SBML Level~3 Package specification for
Distributions, Version~1, \sec{uncertbound-class}.)}

\validRule{distrib-21304}{The attribute \token{distrib:\-inclusive} on
an \UncertBound must have a value of data type \token{boolean}.
(Reference: SBML Level~3 Package specification for Distributions,
Version~1, \sec{uncertbound-class}.)}


\subsubsection*{Rules for \class{ExternalDistribution} object}

\validRule{distrib-21401}{An \ExternalDistribution object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on an \ExternalDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-21402}{An \ExternalDistribution object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on an
\ExternalDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-21403}{An \ExternalDistribution object must have the
required attribute \token{distrib:\-definitionURL}. No other attributes
from the SBML Level 3 Distributions namespaces are permitted on an
\ExternalDistribution object. (Reference: SBML Level~3 Package
specification for Distributions, Version~1,
\sec{externaldistribution-class}.)}

\validRule{distrib-21404}{An \ExternalDistribution object may contain
one and only one instance of the \ListOfExternalParameters element. No
other elements from the SBML Level 3 Distributions namespaces are
permitted on an \ExternalDistribution object. (Reference: SBML Level~3
Package specification for Distributions, Version~1,
\sec{externaldistribution-class}.)}

\validRule{distrib-21405}{The attribute \token{distrib:\-definitionURL}
on an \ExternalDistribution must have a value of data type
\token{string}. (Reference: SBML Level~3 Package specification for
Distributions, Version~1, \sec{externaldistribution-class}.)}

\validRule{distrib-21406}{Apart from the general notes and annotations
subobjects permitted on all SBML objects, a \ListOfExternalParameters
container object may only contain \ExternalParameter objects.
(Reference: SBML Level~3 Package specification for Distributions,
Version~1, \sec{listofexternalparameters-class}.)}

\validRule{distrib-21407}{A \ListOfExternalParameters object may have
the optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \ListOfExternalParameters object.
(Reference: SBML Level~3 Package specification for Distributions,
Version~1, \sec{listofexternalparameters-class}.)}


\subsubsection*{Rules for \class{ExternalParameter} object}

\validRule{distrib-21501}{An \ExternalParameter object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on an \ExternalParameter. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-21502}{An \ExternalParameter object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on an
\ExternalParameter. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-21503}{An \ExternalParameter object must have the
required attribute \token{distrib:\-definitionURL}. No other attributes
from the SBML Level 3 Distributions namespaces are permitted on an
\ExternalParameter object. (Reference: SBML Level~3 Package
specification for Distributions, Version~1,
\sec{externalparameter-class}.)}

\validRule{distrib-21504}{An \ExternalParameter object may contain one
and only one instance of the \ListOfExternalParameters element. No other
elements from the SBML Level 3 Distributions namespaces are permitted on
an \ExternalParameter object. (Reference: SBML Level~3 Package
specification for Distributions, Version~1,
\sec{externalparameter-class}.)}

\validRule{distrib-21505}{The attribute \token{distrib:\-definitionURL}
on an \ExternalParameter must have a value of data type \token{string}.
(Reference: SBML Level~3 Package specification for Distributions,
Version~1, \sec{externalparameter-class}.)}

\validRule{distrib-21506}{Apart from the general notes and annotations
subobjects permitted on all SBML objects, a \ListOfExternalParameters
container object may only contain \ExternalParameter objects.
(Reference: SBML Level~3 Package specification for Distributions,
Version~1, \sec{listofexternalparameters-class}.)}

\validRule{distrib-21507}{A \ListOfExternalParameters object may have
the optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \ListOfExternalParameters object.
(Reference: SBML Level~3 Package specification for Distributions,
Version~1, \sec{listofexternalparameters-class}.)}


\subsubsection*{Rules for \class{NormalDistribution} object}

\validRule{distrib-21601}{A \NormalDistribution object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \NormalDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-21602}{A \NormalDistribution object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on a
\NormalDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-21603}{A \NormalDistribution object must contain one
and only one instance of the UncertValue element, and may contain one
and only one instance of each of the UncertValue and UncertValue
elements. No other elements from the SBML Level 3 Distributions
namespaces are permitted on a \NormalDistribution object. (Reference:
SBML Level~3 Package specification for Distributions, Version~1,
\sec{normaldistribution-class}.)}


\subsubsection*{Rules for \class{UniformDistribution} object}

\validRule{distrib-21701}{An \UniformDistribution object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on an \UniformDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-21702}{An \UniformDistribution object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on an
\UniformDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-21703}{An \UniformDistribution object must contain
one and only one instance of each of the UncertValue and \UncertValue
elements, and may contain one and only one instance of the UncertValue
element. No other elements from the SBML Level 3 Distributions
namespaces are permitted on an \UniformDistribution object. (Reference:
SBML Level~3 Package specification for Distributions, Version~1,
\sec{uniformdistribution-class}.)}


\subsubsection*{Rules for \class{CategoricalDistribution} object}

\validRule{distrib-21801}{A \CategoricalDistribution object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \CategoricalDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-21802}{A \CategoricalDistribution object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on a
\CategoricalDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-21803}{A \CategoricalDistribution object must contain
one and only one instance of the \ListOfCategories element. No other
elements from the SBML Level 3 Distributions namespaces are permitted on
a \CategoricalDistribution object. (Reference: SBML Level~3 Package
specification for Distributions, Version~1,
\sec{categoricaldistribution-class}.)}

\validRule{distrib-21804}{The \ListOfCategories subobject on a
\CategoricalDistribution object must not be empty. (Reference: SBML
Level~3 Package specification for Distributions, Version~1,
\sec{categoricaldistribution-class}.)}

\validRule{distrib-21805}{Apart from the general notes and annotations
subobjects permitted on all SBML objects, a \ListOfCategories container
object may only contain \Category objects. (Reference: SBML Level~3
Package specification for Distributions, Version~1,
\sec{listofcategories-class}.)}

\validRule{distrib-21806}{A \ListOfCategories object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \ListOfCategories object. (Reference: SBML
Level~3 Package specification for Distributions, Version~1,
\sec{listofcategories-class}.)}


\subsubsection*{Rules for \class{Category} object}

\validRule{distrib-21901}{A \Category object may have the optional SBML
Level~3 Core attributes \token{metaid} and \token{sboTerm}. No other
attributes from the SBML Level 3 Core namespaces are permitted on a
\Category. (Reference: SBML Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-21902}{A \Category object may have the optional SBML
Level~3 Core subobjects for notes and annotations. No other elements
from the SBML Level 3 Core namespaces are permitted on a \Category.
(Reference: SBML Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-21903}{A \Category object may have the optional
attribute \token{distrib:\-rank}. No other attributes from the SBML
Level 3 Distributions namespaces are permitted on a \Category object.
(Reference: SBML Level~3 Package specification for Distributions,
Version~1, \sec{category-class}.)}

\validRule{distrib-21904}{A \Category object must contain one and only
one instance of the UncertValue element, and may contain one and only
one instance of the UncertValue element. No other elements from the SBML
Level 3 Distributions namespaces are permitted on a \Category object.
(Reference: SBML Level~3 Package specification for Distributions,
Version~1, \sec{category-class}.)}

\validRule{distrib-21905}{The attribute \token{distrib:\-rank} on a
\Category must have a value of data type \token{integer}. (Reference:
SBML Level~3 Package specification for Distributions, Version~1,
\sec{category-class}.)}


\subsubsection*{Rules for \class{BernoulliDistribution} object}

\validRule{distrib-22001}{A \BernoulliDistribution object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \BernoulliDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-22002}{A \BernoulliDistribution object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on a
\BernoulliDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-22003}{A \BernoulliDistribution object must contain
one and only one instance of the UncertValue element. No other elements
from the SBML Level 3 Distributions namespaces are permitted on a
\BernoulliDistribution object. (Reference: SBML Level~3 Package
specification for Distributions, Version~1,
\sec{bernoullidistribution-class}.)}


\subsubsection*{Rules for \class{BetaDistribution} object}

\validRule{distrib-22101}{A \BetaDistribution object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \BetaDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-22102}{A \BetaDistribution object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on a
\BetaDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-22103}{A \BetaDistribution object must contain one
and only one instance of each of the UncertValue and \UncertValue
elements. No other elements from the SBML Level 3 Distributions
namespaces are permitted on a \BetaDistribution object. (Reference: SBML
Level~3 Package specification for Distributions, Version~1,
\sec{betadistribution-class}.)}


\subsubsection*{Rules for \class{BinomialDistribution} object}

\validRule{distrib-22201}{A \BinomialDistribution object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \BinomialDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-22202}{A \BinomialDistribution object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on a
\BinomialDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-22203}{A \BinomialDistribution object must contain
one and only one instance of each of the UncertValue and \UncertValue
elements. No other elements from the SBML Level 3 Distributions
namespaces are permitted on a \BinomialDistribution object. (Reference:
SBML Level~3 Package specification for Distributions, Version~1,
\sec{binomialdistribution-class}.)}


\subsubsection*{Rules for \class{CauchyDistribution} object}

\validRule{distrib-22301}{A \CauchyDistribution object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \CauchyDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-22302}{A \CauchyDistribution object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on a
\CauchyDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-22303}{A \CauchyDistribution object must contain one
and only one instance of each of the UncertValue and \UncertValue
elements. No other elements from the SBML Level 3 Distributions
namespaces are permitted on a \CauchyDistribution object. (Reference:
SBML Level~3 Package specification for Distributions, Version~1,
\sec{cauchydistribution-class}.)}


\subsubsection*{Rules for \class{ChiSquareDistribution} object}

\validRule{distrib-22401}{A \ChiSquareDistribution object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \ChiSquareDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-22402}{A \ChiSquareDistribution object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on a
\ChiSquareDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-22403}{A \ChiSquareDistribution object must contain
one and only one instance of the UncertValue element. No other elements
from the SBML Level 3 Distributions namespaces are permitted on a
\ChiSquareDistribution object. (Reference: SBML Level~3 Package
specification for Distributions, Version~1,
\sec{chisquaredistribution-class}.)}


\subsubsection*{Rules for \class{ExponentialDistribution} object}

\validRule{distrib-22501}{An \ExponentialDistribution object may have
the optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on an \ExponentialDistribution. (Reference:
SBML Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-22502}{An \ExponentialDistribution object may have
the optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on an
\ExponentialDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-22503}{An \ExponentialDistribution object must
contain one and only one instance of the UncertValue element. No other
elements from the SBML Level 3 Distributions namespaces are permitted on
an \ExponentialDistribution object. (Reference: SBML Level~3 Package
specification for Distributions, Version~1,
\sec{exponentialdistribution-class}.)}


\subsubsection*{Rules for \class{FDistribution} object}

\validRule{distrib-22601}{A \FDistribution object may have the optional
SBML Level~3 Core attributes \token{metaid} and \token{sboTerm}. No
other attributes from the SBML Level 3 Core namespaces are permitted on
a \FDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-22602}{A \FDistribution object may have the optional
SBML Level~3 Core subobjects for notes and annotations. No other
elements from the SBML Level 3 Core namespaces are permitted on a
\FDistribution. (Reference: SBML Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-22603}{A \FDistribution object must contain one and
only one instance of each of the UncertValue and \UncertValue elements.
No other elements from the SBML Level 3 Distributions namespaces are
permitted on a \FDistribution object. (Reference: SBML Level~3 Package
specification for Distributions, Version~1, \sec{fdistribution-class}.)}


\subsubsection*{Rules for \class{GammaDistribution} object}

\validRule{distrib-22701}{A \GammaDistribution object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \GammaDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-22702}{A \GammaDistribution object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on a
\GammaDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-22703}{A \GammaDistribution object must contain one
and only one instance of each of the UncertValue and \UncertValue
elements. No other elements from the SBML Level 3 Distributions
namespaces are permitted on a \GammaDistribution object. (Reference:
SBML Level~3 Package specification for Distributions, Version~1,
\sec{gammadistribution-class}.)}


\subsubsection*{Rules for \class{GeometricDistribution} object}

\validRule{distrib-22801}{A \GeometricDistribution object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \GeometricDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-22802}{A \GeometricDistribution object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on a
\GeometricDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-22803}{A \GeometricDistribution object must contain
one and only one instance of the UncertValue element. No other elements
from the SBML Level 3 Distributions namespaces are permitted on a
\GeometricDistribution object. (Reference: SBML Level~3 Package
specification for Distributions, Version~1,
\sec{geometricdistribution-class}.)}


\subsubsection*{Rules for \class{HypergeometricDistribution} object}

\validRule{distrib-22901}{A \HypergeometricDistribution object may have
the optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \HypergeometricDistribution. (Reference:
SBML Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-22902}{A \HypergeometricDistribution object may have
the optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on a
\HypergeometricDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-22903}{A \HypergeometricDistribution object must
contain one and only one instance of each of the UncertValue,
UncertValue and \UncertValue elements. No other elements from the SBML
Level 3 Distributions namespaces are permitted on a
\HypergeometricDistribution object. (Reference: SBML Level~3 Package
specification for Distributions, Version~1,
\sec{hypergeometricdistribution-class}.)}


\subsubsection*{Rules for \class{InverseGammaDistribution} object}

\validRule{distrib-23001}{An \InverseGammaDistribution object may have
the optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on an \InverseGammaDistribution. (Reference:
SBML Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-23002}{An \InverseGammaDistribution object may have
the optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on an
\InverseGammaDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-23003}{An \InverseGammaDistribution object must
contain one and only one instance of each of the UncertValue and
\UncertValue elements. No other elements from the SBML Level 3
Distributions namespaces are permitted on an \InverseGammaDistribution
object. (Reference: SBML Level~3 Package specification for
Distributions, Version~1, \sec{inversegammadistribution-class}.)}


\subsubsection*{Rules for \class{LaplaceDistribution} object}

\validRule{distrib-23101}{A \LaplaceDistribution object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \LaplaceDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-23102}{A \LaplaceDistribution object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on a
\LaplaceDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-23103}{A \LaplaceDistribution object must contain one
and only one instance of each of the UncertValue and \UncertValue
elements. No other elements from the SBML Level 3 Distributions
namespaces are permitted on a \LaplaceDistribution object. (Reference:
SBML Level~3 Package specification for Distributions, Version~1,
\sec{laplacedistribution-class}.)}


\subsubsection*{Rules for \class{LogNormalDistribution} object}

\validRule{distrib-23201}{A \LogNormalDistribution object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \LogNormalDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-23202}{A \LogNormalDistribution object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on a
\LogNormalDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-23203}{A \LogNormalDistribution object must contain
one and only one instance of each of the UncertValue and \UncertValue
elements. No other elements from the SBML Level 3 Distributions
namespaces are permitted on a \LogNormalDistribution object. (Reference:
SBML Level~3 Package specification for Distributions, Version~1,
\sec{lognormaldistribution-class}.)}


\subsubsection*{Rules for \class{LogisticDistribution} object}

\validRule{distrib-23301}{A \LogisticDistribution object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \LogisticDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-23302}{A \LogisticDistribution object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on a
\LogisticDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-23303}{A \LogisticDistribution object must contain
one and only one instance of each of the UncertValue and \UncertValue
elements. No other elements from the SBML Level 3 Distributions
namespaces are permitted on a \LogisticDistribution object. (Reference:
SBML Level~3 Package specification for Distributions, Version~1,
\sec{logisticdistribution-class}.)}


\subsubsection*{Rules for \class{NegativeBinomialDistribution} object}

\validRule{distrib-23401}{A \NegativeBinomialDistribution object may
have the optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \NegativeBinomialDistribution. (Reference:
SBML Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-23402}{A \NegativeBinomialDistribution object may
have the optional SBML Level~3 Core subobjects for notes and
annotations. No other elements from the SBML Level 3 Core namespaces are
permitted on a \NegativeBinomialDistribution. (Reference: SBML Level~3
Version~1 Core, Section~3.2.)}

\validRule{distrib-23403}{A \NegativeBinomialDistribution object must
contain one and only one instance of each of the UncertValue and
\UncertValue elements. No other elements from the SBML Level 3
Distributions namespaces are permitted on a
\NegativeBinomialDistribution object. (Reference: SBML Level~3 Package
specification for Distributions, Version~1,
\sec{negativebinomialdistribution-class}.)}


\subsubsection*{Rules for \class{ParetoDistribution} object}

\validRule{distrib-23501}{A \ParetoDistribution object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \ParetoDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-23502}{A \ParetoDistribution object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on a
\ParetoDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-23503}{A \ParetoDistribution object must contain one
and only one instance of each of the UncertValue and \UncertValue
elements. No other elements from the SBML Level 3 Distributions
namespaces are permitted on a \ParetoDistribution object. (Reference:
SBML Level~3 Package specification for Distributions, Version~1,
\sec{paretodistribution-class}.)}


\subsubsection*{Rules for \class{PoissonDistribution} object}

\validRule{distrib-23601}{A \PoissonDistribution object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \PoissonDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-23602}{A \PoissonDistribution object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on a
\PoissonDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-23603}{A \PoissonDistribution object must contain one
and only one instance of the UncertValue element. No other elements from
the SBML Level 3 Distributions namespaces are permitted on a
\PoissonDistribution object. (Reference: SBML Level~3 Package
specification for Distributions, Version~1,
\sec{poissondistribution-class}.)}


\subsubsection*{Rules for \class{RayleighDistribution} object}

\validRule{distrib-23701}{A \RayleighDistribution object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \RayleighDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-23702}{A \RayleighDistribution object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on a
\RayleighDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-23703}{A \RayleighDistribution object must contain
one and only one instance of the UncertValue element. No other elements
from the SBML Level 3 Distributions namespaces are permitted on a
\RayleighDistribution object. (Reference: SBML Level~3 Package
specification for Distributions, Version~1,
\sec{rayleighdistribution-class}.)}


\subsubsection*{Rules for \class{StudentTDistribution} object}

\validRule{distrib-23801}{A \StudentTDistribution object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \StudentTDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-23802}{A \StudentTDistribution object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on a
\StudentTDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-23803}{A \StudentTDistribution object must contain
one and only one instance of each of the UncertValue, UncertValue and
\UncertValue elements. No other elements from the SBML Level 3
Distributions namespaces are permitted on a \StudentTDistribution
object. (Reference: SBML Level~3 Package specification for
Distributions, Version~1, \sec{studenttdistribution-class}.)}


\subsubsection*{Rules for \class{WeibullDistribution} object}

\validRule{distrib-23901}{A \WeibullDistribution object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \WeibullDistribution. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-23902}{A \WeibullDistribution object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on a
\WeibullDistribution. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-23903}{A \WeibullDistribution object must contain one
and only one instance of each of the UncertValue and \UncertValue
elements. No other elements from the SBML Level 3 Distributions
namespaces are permitted on a \WeibullDistribution object. (Reference:
SBML Level~3 Package specification for Distributions, Version~1,
\sec{weibulldistribution-class}.)}


\subsubsection*{Rules for \class{Uncertainty} object}

\validRule{distrib-24001}{An \Uncertainty object may have the optional
SBML Level~3 Core attributes \token{metaid} and \token{sboTerm}. No
other attributes from the SBML Level 3 Core namespaces are permitted on
an \Uncertainty. (Reference: SBML Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-24002}{An \Uncertainty object may have the optional
SBML Level~3 Core subobjects for notes and annotations. No other
elements from the SBML Level 3 Core namespaces are permitted on an
\Uncertainty. (Reference: SBML Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-24003}{An \Uncertainty object may contain one and
only one instance of each of the UncertStatistics and Distribution
elements. No other elements from the SBML Level 3 Distributions
namespaces are permitted on an \Uncertainty object. (Reference: SBML
Level~3 Package specification for Distributions, Version~1,
\sec{uncertainty-class}.)}


\subsubsection*{Rules for \class{UncertStatistics} object}

\validRule{distrib-24101}{An \UncertStatistics object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on an \UncertStatistics. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-24102}{An \UncertStatistics object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on an
\UncertStatistics. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-24103}{An \UncertStatistics object may contain one
and only one instance of each of the UncertValue, UncertValue,
UncertValue, UncertValue, UncertValue, UncertValue, UncertValue,
UncertValue, UncertStatisticSpan, UncertStatisticSpan,
UncertStatisticSpan, UncertStatisticSpan and \ListOfExternalParameters
elements. No other elements from the SBML Level 3 Distributions
namespaces are permitted on an \UncertStatistics object. (Reference:
SBML Level~3 Package specification for Distributions, Version~1,
\sec{uncertstatistics-class}.)}

\validRule{distrib-24104}{Apart from the general notes and annotations
subobjects permitted on all SBML objects, a \ListOfExternalParameters
container object may only contain \ExternalParameter objects.
(Reference: SBML Level~3 Package specification for Distributions,
Version~1, \sec{listofexternalparameters-class}.)}

\validRule{distrib-24105}{A \ListOfExternalParameters object may have
the optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on a \ListOfExternalParameters object.
(Reference: SBML Level~3 Package specification for Distributions,
Version~1, \sec{listofexternalparameters-class}.)}


\subsubsection*{Rules for \class{UncertStatisticSpan} object}

\validRule{distrib-24201}{An \UncertStatisticSpan object may have the
optional SBML Level~3 Core attributes \token{metaid} and
\token{sboTerm}. No other attributes from the SBML Level 3 Core
namespaces are permitted on an \UncertStatisticSpan. (Reference: SBML
Level~3 Version~1 Core, Section~3.2.)}

\validRule{distrib-24202}{An \UncertStatisticSpan object may have the
optional SBML Level~3 Core subobjects for notes and annotations. No
other elements from the SBML Level 3 Core namespaces are permitted on an
\UncertStatisticSpan. (Reference: SBML Level~3 Version~1 Core,
Section~3.2.)}

\validRule{distrib-24203}{An \UncertStatisticSpan object may have the
optional attributes \token{distrib:\-varLower},
\token{distrib:\-valueLower}, \token{distrib:\-varUpper} and
\token{distrib:\-valueUpper}. No other attributes from the SBML Level 3
Distributions namespaces are permitted on an \UncertStatisticSpan
object. (Reference: SBML Level~3 Package specification for
Distributions, Version~1, \sec{uncertstatisticspan-class}.)}

\validRule{distrib-24204}{The value of the attribute
\token{distrib:\-varLower} of an \UncertStatisticSpan object must be the
identifier of an existing \SBase object defined in the enclosing \Model
object. (Reference: SBML Level~3 Package specification for
Distributions, Version~1, \sec{uncertstatisticspan-class}.)}

\validRule{distrib-24205}{The attribute \token{distrib:\-valueLower} on
an \UncertStatisticSpan must have a value of data type \token{double}.
(Reference: SBML Level~3 Package specification for Distributions,
Version~1, \sec{uncertstatisticspan-class}.)}

\validRule{distrib-24206}{The value of the attribute
\token{distrib:\-varUpper} of an \UncertStatisticSpan object must be the
identifier of an existing \SBase object defined in the enclosing \Model
object. (Reference: SBML Level~3 Package specification for
Distributions, Version~1, \sec{uncertstatisticspan-class}.)}

\validRule{distrib-24207}{The attribute \token{distrib:\-valueUpper} on
an \UncertStatisticSpan must have a value of data type \token{double}.
(Reference: SBML Level~3 Package specification for Distributions,
Version~1, \sec{uncertstatisticspan-class}.)}


