% -*- TeX-master: "main" -*-

\section{Introduction}
\label{intro}

The SBML Level~3 Required Elements package is a small package that allows models to declare exactly which components of a model have the meaning of their mathematical semantics changed by another package or by custom annotations.  It does this simply by defining an optional child of \SBase which can be attached to any component of a model with associated mathematics.  The rest of this specification explains the package constructs and how they can be used.


\subsection{Proposal corresponding to this package specification}

This specification for Required Elements in SBML Level~3 Version~1 is based on the draft located at this URL:

\begin{center}
  \vspace*{1ex}\small
  \url{http://sbml.svn.sf.net/svnroot/sbml/trunk/specifications/sbml-level-3/version-1/req/specification/}
  \vspace*{1ex}
\end{center}

The tracking number in the SBML issue tracking system~\citep{tracker} for Required Elements package activities is 3594695.  The version of the proposal used as the starting point for this specification is the version of August, 2011~\citep{req_proposal_2011}.


\subsection{Package dependencies}

The Required Elements package has no dependencies on other SBML Level~3 packages.


\subsection{Document conventions}
\label{conventions}

UML~1.0 (Unified Modeling Language; \citealt{eriksson:1998, oestereich:1999}) notation is used in this document to define the constructs provided by this package.  Colors in the diagrams carry the following additional information for the benefit of those viewing the document on media that can display color:

\begin{itemize}

\item[\raisebox{2.75pt}{\colorbox{black}{\rule{0.8pt}{0.8pt}}}]
  \emph{Black}: Items colored black in the UML diagrams are components
  taken unchanged from their definition in the SBML Level~3 Core
  specification document.

\item[\raisebox{2.75pt}{\colorbox{mediumgreen}{\rule{0.8pt}{0.8pt}}}]
  \emph{\textcolor{mediumgreen}{Green}}: Items colored green are
  components that exist in SBML Level~3 Core, but are extended by this
  package.  Class boxes are also drawn with dashed lines to further
  distinguish them.

\item[\raisebox{2.75pt}{\colorbox{darkblue}{\rule{0.8pt}{0.8pt}}}]
  \emph{\textcolor{darkblue}{Blue}}: Items colored blue are new
  components introduced in this package specification.  They have no
  equivalent in the SBML Level~3 Core specification.

\end{itemize}

The following typographical conventions distinguish the names of objects and data types from other entities; these conventions are identical to the conventions used in the SBML Level~3 Core specification document:

\begin{description}
  
\item \class{Class}: Names of ordinary (concrete) classes begin with a capital letter and are printed in an upright, bold, sans-serif typeface.  In electronic document formats, the class names are also hyperlinked to their definitions in this specification document.  (However, class names are not hyperlinked if they are for classes that are unchanged from their definitions in their original specification.)

\item \token{SomeThing}, \token{otherThing}: Attributes of classes, data type names, literal XML, and tokens \emph{other} than SBML class names, are printed in an upright typewriter typeface.

\end{description}

For other matters involving the use of UML and XML, this document follows the conventions used in the SBML Level~3 Core specification document.
